---
title: "Paashi_Paleo_Models_Pub"
author: "Robin_ML"
date: "2026-09-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages and Raw Data
```{r loading packages/data}

library(tidyverse)
library(tidybayes)
library(ggplot2)
library(ggthemes)
library(brms)
library(bayesplot)
library(loo)
library(bridgesampling)
library(patchwork)
library(scales)
library(ggrepel)
library(maps)
library(ggnewscale)

clim <- read.csv("Raw Data/ElevSSTs.csv")

```

## Publication Theme and Palette (assisted by Claude Code)
```{r theme_and_palette}
# Colorblind-safe, grayscale-distinguishable palette
col_model_line   <- "#1F4E79"   # dark blue — model trend line
col_model_ribbon <- "#A8C8E8"   # light blue — prediction interval
col_model_inner  <- "#4682B4"   # steel blue — fitted CI
col_observed     <- "gray20"    # near-black — observed data
col_arch         <- "#B85C38"   # terra cotta — archaeological reconstruction
col_seki         <- "#D4930D"   # amber — Seki SST (California Margin)
col_stott        <- "#1B998B"   # teal — Stott SST (W. Equatorial Pacific)
col_davis5       <- "#5B7B9A"   # slate blue — Davis5 SST (Sea of Okhotsk)
col_pahnke       <- "#A65A6E"   # muted rose — Pahnke SST (E. Tropical Pacific, Panama Basin)
col_train_bg     <- "#FFF8F0"   # faint warm — calibration window background
col_pleis_line   <- "gray25"    # Pleistocene-Holocene boundary
col_pdo_line     <- "#B85C38"
col_wb_draw      <- "#4682B4"

# Shared theme for JAS Reports (Elsevier) — base set to match ~10pt body type. With the existing rel() multipliers, axis text lands at 9pt and annotations (size = 3.0) at ~8.5pt, preserving the title > body > axis > annotation hierarchy.
theme_paashi <- function(base_size = 10) {
  theme_minimal(base_size = base_size, base_family = "sans") %+replace%
    theme(
      plot.title       = element_text(size = rel(1.15), face = "bold",
                                      hjust = 0, margin = margin(b = 4)),
      plot.subtitle    = element_text(size = rel(0.95), color = "gray30",
                                      hjust = 0, margin = margin(b = 6)),
      axis.title       = element_text(size = rel(1.0)),
      axis.text        = element_text(size = rel(0.9), color = "gray20"),
      strip.text       = element_text(size = rel(1.0), face = "bold"),
      legend.text      = element_text(size = rel(0.9)),
      legend.title     = element_text(size = rel(0.95), face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
      plot.margin      = margin(8, 8, 8, 8),
      legend.position  = "bottom",
      legend.key.size  = unit(0.4, "cm")
    )
}

# Save figures at JAS dimensions
save_fig <- function(plot, filename,
                     width_mm = 190, height_mm = 120,
                     dpi = 500, device = "tiff") {
  ggsave(
    filename = file.path("Figures", filename),
    plot     = plot,
    width    = width_mm, height = height_mm,
    units    = "mm", dpi = dpi, device = device,
    compression = if (device == "tiff") "lzw" else NULL
  )
}

dir.create("Figures", showWarnings = FALSE)

#we report ages in cal yr BP (present = AD 1950). Our data are stored as years before 2005 (Year_B2005), so plots subtract BP_OFFSET to convert.
BP_OFFSET <- 55
```

## Checking Paleoclimate Data
```{r plotting}

##drop NAs and plot candidate predictors
pl <- function(v, main) {
  ok <- !is.na(clim[[v]])
  plot(clim$Year_B2005[ok] - BP_OFFSET, clim[[v]][ok], main = main, type = "o",
       pch = 20, cex = 0.5, xlim = c(0, 20000),
       xlab = "cal yr BP", ylab = "SST (C)")
}

par(mfrow = c(4, 3))
pl("Barron", "Barron (E Pacific, 15kya, ~120yr res)")
pl("Davis2", "Davis-2 (E Pacific, 15kya, ~100yr res)")
pl("Davis3", "Davis-3 (Okhotsk, 24kya, ~90yr res)")
pl("Davis4", "Davis-4 (Okhotsk, 15kya, ~480yr res, d18O)")
pl("Davis5", "Davis-5 (Okhotsk, 25kya, ~410yr res)")
pl("Davis6", "Davis-6 (Okhotsk, 26kya, ~540yr res, d18O)")
pl("Davis8", "Davis-8 (Okhotsk, 15kya, ~20yr res)")
pl("Isono", "Isono (Japan, 16kya, ~70yr res)")
pl("Pahnke", "Pahnke (E Trop Pacific, 27kya, ~370yr res)")
pl("Seki", "Seki (California Margin, 29kya, ~260yr res)")
pl("Stott", "Stott (W Trop Pacific, 20kya, ~65yr res)")
pl("Yamamoto", "Yamamoto (California Margin, 29kya, ~1430yr res) EXCLUDED")
par(mfrow = c(1, 1))
#Yamamoto can be excluded because it has only one observation in the training period, so fitting a model to it is just fitting to a flat line, essentially no information. Also, its median sampling interval is ~1,430 years, which fails our resolution criterion of 1,000 years or finer.


##Examine calibration data
fig_elev_hist <- ggplot(clim, aes(x = elev)) +
  geom_histogram(bins = 300, fill = col_model_ribbon, color = "white", linewidth = 0.1) +
  labs(x = "Lake Elevation (masl)", y = "Count",
       title = "Distribution of Observed Lake Elevation (Calibration Period)",
       subtitle = paste0("Mean = ", round(mean(clim$elev, na.rm = TRUE), 1),
                         " masl; SD = ", round(sd(clim$elev, na.rm = TRUE), 2),
                         " m")) +
  theme_paashi()
fig_elev_hist
save_fig(fig_elev_hist, "Figure_S1_Elevation_Histogram.tiff", height_mm = 90)
#This gives a normal distribution, so no further action needed; we can use family = gaussian. The distribution may in fact be bimodal with two very close peaks near the center-- this may represent "drought" and "flood" regimes, as truly average rainfall is rarity in California according to oral traditions. However, these peaks are close enough to treat as a single normal distribution; bimodal modeling produces bad results

```

## Interpolating SST Records and Adding Uncertainty
```{r interpolation and uncertainty}
##We need to interpolate to annual resolution so ARMA can work later. The reviewers also requested that we add uncertainty from measurement error, age-depth error, and interpolation. We are going to follow McKay's workflow by producing randomized draws, which will allow us to propagate uncertainty later by fitting models from each draw, making predictions from each of those models, and learning from all of the posteriors.

#measurement: measurement error is not reported for every SST record, so each measurement wobbles by a standard MEAS_SD

#chronology: time series error is not reported for all SST records, so each draw shifts each record's whole timescale randomly according to AGE_SD

#interpolation: we create draws between each point that wander on a "brownian bridge" (see comment below); we keep a linear interpolation as the the mean of these draws. The draws inform uncertainty around the interpolation, with the most uncertainty halfway between measurements.

##Draws are saved to an rds so models can later access each draw to propagate uncertainty

##Keep only the records that meet both predictor criteria: resolution of 1,000 years or finer (drops Yamamoto) and a span of at least 20,000 years (drops Isono, Barron, Davis2, Davis8 and Davis4).
##Keep the first five in this order: draws are generated one record at a time, so reordering them would change the draws for the four final predictors.
SCREENED <- c("Seki", "Stott", "Davis5", "Davis3", "Pahnke",
              "Davis6")

N_DRAWS  <- 50    #number of draws. this is a factor in ESS for uncertainty propagation; we use 50 because it gives relatively reliable results without heavy compute or washing out trends in reconstruction
MEAS_SD  <- 0.2   #SD of degrees C to wobble each measurement. No record in our pool reports a measurement SD. nearest published figures is  +/-0.2 C (Harada et al. 2012, for core MR06-04-PC04 = our Davis8). 0.2 SD leaves a small buffer around that reported duplicate difference.
AGE_SD   <- 150   #SD of years to shift each whole record for each draw; likewise time-depth uncertainty is not reported by our SST sources.
set.seed(2025)

##points_uncertainty shifts the time scale of the whole record (like an age-depth model); wobbles temperature at each point (like measurement error)
points_uncertainty <- function(pts) {
  data.frame(
    age = pts$age + rnorm(1, 0, AGE_SD), #slide the record's timescale by a random amount
    sst = pts$sst + rnorm(nrow(pts), 0, MEAS_SD) #wobble each measured temperature
  )
}

##find_wander_rate measures how fast each original record's temperature moves, as variance gained per year. Also called tau^2. We need it to build the plausible paths between the randomized records (next step). for each neighboring pair of measurements, square the temperature change and divide by the years between them, then take the median across all pairs. That is the typical squared movement per year. multiply by elapsed time and take sqrt() to get back to the standard deviation.
find_wander_rate <- function(pts) {
  median(diff(pts$sst)^2 / diff(pts$age))
}

##sim_path creates a plausible annual path through each draw's measurements. The random path between measurements is a "Brownian bridge": a random path between measurements that always lands exactly on the next known measurement. It works additively, so with many draws it creates a normal distribution of possible temps at each year. Across many draws, uncertainty will be zero at each known measurement and largest halfway between them. Because we also wobble measurements with meas_sd and age_sd for each draw, the aggregate will also have measurement uncertainty and age uncertainty at each point.
sim_path <- function(years, age, sst, wander_rate) {
  #put the measurements in time order, oldest to youngest
  o <- order(age); age <- age[o]; sst <- sst[o]
  #draw a straight line between measurements. This is the path we will add random wander to. rule = 2 means extrapolation beyond the data is just flat
  path <- approx(age, sst, xout = years, rule = 2)$y
  dev  <- numeric(length(years)) #Deviation. We initiate at 0, and the following steps will generate random wander to produce deviation
  ##BETWEEN measurements: build a bridge across each gap
  for (i in seq_len(length(age) - 1)) {
    #the years inside the gap
    j <- which(years > age[i] & years < age[i + 1])
    if (!length(j)) next
    #the time axis for the bridge. The two known measurements are the start and end points
    t_all <- c(age[i], years[j], age[i + 1])
    #random walk from start point time to end point time. Each step nudges the value randomly from the last value, according to wander rate / tau2, so the path is a continuous sequence. That creates additive variance; sqrt() converts that variance into standard deviation for rnorm.
    walk <- cumsum(c(0, rnorm(length(t_all) - 1, 0, sqrt(wander_rate * diff(t_all)))))
    #How far through the gap each step is, 0 to 1
    frac <- (t_all - t_all[1]) / (t_all[length(t_all)] - t_all[1])
    #Subtract the known endpoint, scaled by fraction of the way through the walk. Makes sure the walk always ends up at the endpoint, with the most variance right in the middle, creating the "bridge."
    bridge <- walk - walk[length(walk)] * frac
    #keep only the in-between years. The measurements at each end are already "known"
    dev[j] <- bridge[-c(1, length(bridge))]
  }

  ##OUTSIDE the measurements there is no end point, so the walk is left free and the uncertainty keeps growing with distance from the nearest measurement.
  #years older than the record's oldest measurement
  lo <- which(years < age[1])
  if (length(lo))
    dev[rev(lo)] <- cumsum(rnorm(length(lo), 0,
                    sqrt(wander_rate * abs(diff(c(age[1], rev(years[lo])))))))
  #years younger than the record's youngest measurement
  hi <- which(years > age[length(age)])
  if (length(hi))
    dev[hi] <- cumsum(rnorm(length(hi), 0,
               sqrt(wander_rate * diff(c(age[length(age)], years[hi])))))

  ##the finished full path of the record: straight mean line plus wandering bridges
  path + dev
}


##interp_record combines the steps above into a loop to build N_DRAWS versions of each record; each draw is a different plausible version of that record based on our age-depth and measurement uncertainty.
interp_record <- function(v) {
  #pull out the years where this record has a measurement
  pts   <- na.omit(data.frame(age = clim$Year_B2005, sst = clim[[v]]))
  pts   <- pts[order(pts$age), ]
  years <- clim$Year_B2005
  #measure this record's wander rate / tau2 from the original record
  wander <- find_wander_rate(pts)
  #build the draws
  draws <- replicate(N_DRAWS, {
    p <- points_uncertainty(pts) #shift and wobble this record's measurements
    sim_path(years, p$age, p$sst, wander) #generate a random path through those shifted, wobbled measurements
  })

  list(mean  = approx(pts$age, pts$sst, xout = years, rule = 2)$y,  #linear interpolation between each point acts as the mean of random draws
       se    = apply(draws, 1, sd),
       draws = draws)
}

#run it for every screened record. Building the draws takes a while, so they are saved and reloaded on later runs
dir.create("Processed Data", showWarnings = FALSE)
interp_file <- sprintf("Processed Data/SST_draws_n%d_meas%s_age%s.rds",
                       N_DRAWS, MEAS_SD, AGE_SD)
if (file.exists(interp_file)) {
  message("reusing ", interp_file)
  interp <- readRDS(interp_file)
  #the record list is not in the file name, so check it here
  if (!identical(names(interp), SCREENED))
    stop("saved draws are for a different record list; delete ", interp_file)
} else {
  interp <- setNames(lapply(SCREENED, interp_record), SCREENED)
  saveRDS(interp, interp_file)
}

#write the interpolated mean and its spread into clim, one pair of columns per record
for (v in SCREENED) {
  clim[[paste0(v, "_Int")]] <- interp[[v]]$mean
  clim[[paste0(v, "_SE")]]  <- interp[[v]]$se
}

##We can center the interpolated predictors so that our intercepts represent expected lake level at mean SST. This also allows the truncated priors requested by reviewers later on. Save the means so the uncertainty draws can also be centered easily.
sst_means <- setNames(
  sapply(SCREENED, function(v) mean(clim[[paste0(v, "_Int")]])),
  SCREENED)
for (v in SCREENED) {
  clim[[paste0(v, "_Int")]] <- clim[[paste0(v, "_Int")]] - sst_means[v]
}

##Visually check the uncertainty. Solid line is the interpolated mean, dashed lines are +/- 2 SE across the draws, red dots are the real measurements. The band should pinch at every red dot and bulge between them.
par(mfrow = c(3, 2))
for (v in SCREENED) {
  z <- interp[[v]]; a <- clim$Year_B2005 - BP_OFFSET
  plot(a, z$mean, type = "l", main = v, xlim = c(0, 20000),
       ylim = range(c(z$mean - 2 * z$se, z$mean + 2 * z$se), na.rm = TRUE),
       xlab = "cal yr BP", ylab = "SST (C)")
  lines(a, z$mean - 2 * z$se, lty = 2)
  lines(a, z$mean + 2 * z$se, lty = 2)
  points(a, clim[[v]], pch = 20, cex = 0.5, col = "red")
}
par(mfrow = c(1, 1))

#the draws are already saved as Processed Data/SST_draws_*.rds

#add interpolated values to raw data file for later
write.csv(clim, "Raw Data/ElevSSTs.csv", row.names = FALSE)

```

## Define Explicit Priors
```{r define_priors}
## Inspect what brms would choose as defaults for our final model
default_priors <- get_prior(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int + arma(p = 1, q = 1, time = Year_B2005),
  data = clim,
  family = gaussian())

print(default_priors)

## Define weakly informative priors in case others have different versions of R or BRMS
# Intercept: Normal(63, 5) truncated to [55, 70] -- centered on mean observed lake elevation (~63 MaSL). With centered predictors, the intercept is the predicted lake level at mean SST values. Truncated at physical basin limits (55 = modern basin floor, 70 = highest geological shoreline) per R1 suggestion.
# Slopes: Normal(0, 5) -- 1-degree SST change shifts lake by <10m (2 SD); generous since entire observed elev range is ~9m
# Sigma: half-Student-t(3, 0, 3) -- standard weakly informative for residual SD
# AR/MA: brms defaults to uniform(-1, 1)

# Base priors for models without autocorrelation terms
paashi_priors <- c(
  prior(normal(63, 5), class = "Intercept", lb = 55, ub = 70),
  prior(normal(0, 5), class = "b"),
  prior(student_t(3, 0, 3), class = "sigma"))

# Extended priors for ARMA models (ar/ma must be explicit for prior predictive checks)
paashi_priors_arma <- c(
  paashi_priors,
  prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1),
  prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1))

```

## Prior Predictive Check
```{r prior_predictive_check}
# Fit model with sample_prior = "only" to draw from priors without seeing data to verify our priors are weakly informative but plausible
mld_prior_only <- brm(
  elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int +
    arma(p = 1, q = 1, time = Year_B2005),
  data = clim,
  family = gaussian(),
  prior = paashi_priors_arma,
  sample_prior = "only",
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2025)

# Simulated data should include the observed range but not be too close, as these priors are weakly informative
fig_prior_pp <- pp_check(mld_prior_only, ndraws = 100, adjust = 10) +
  labs(title = "Prior Predictive Check",
       subtitle = "Dark = observed; light = simulated from prior only",
       x = "Lake Elevation (masl)", y = "Density") +
  coord_cartesian(xlim = c(47, 77)) +
  theme_paashi()
fig_prior_pp
save_fig(fig_prior_pp, "Figure_S2_Prior_Predictive.tiff", height_mm = 90)
## looks good!

```

## Create and Test Models
```{r screening}
##Fit each SST record as a single predictor, rank by LOO

dir.create("Models", showWarnings = FALSE)
CAL <- which(!is.na(clim$elev))

# Compile once, then update with new data for each record
base_dat <- data.frame(elev = clim$elev[CAL], sst = clim[[paste0(SCREENED[1], "_Int")]][CAL])
base_fit <- brm(elev ~ sst, data = base_dat, family = gaussian(),
                prior = paashi_priors, chains = 4, iter = 3000, warmup = 1000,
                seed = 2025, refresh = 0)

screen <- list()
for (v in SCREENED) {
  f_out <- file.path("Models", paste0("screen_", v, ".rds"))
  if (file.exists(f_out)) {
    screen[[v]] <- readRDS(f_out)
    message("reusing ", v)
  } else {
    message("fitting ", v)
    screen[[v]] <- update(base_fit,
                          newdata = data.frame(elev = clim$elev[CAL],
                                              sst = clim[[paste0(v, "_Int")]][CAL]),
                          recompile = FALSE, refresh = 0)
    saveRDS(screen[[v]], f_out)
  }
}

#compare the six single-predictor models
do.call(loo, c(unname(screen), list(compare = TRUE, model_names = names(screen))))

#Model comparisons:
#       elpd_diff se_diff
#Davis5   0.0       0.0  
#Davis6 -10.0       9.0  
#Davis3 -12.2       6.5  
#Pahnke -14.8       5.5  
#Stott  -19.4       5.6  
#Seki   -20.8       6.2 

```

## Combined Models
```{r combined}
###Now we fit and test several candidate models made up of multiple SST records. We construct these models based on their loo rankings and on the records' plausible mechanical connections to California precipitation, according to existing literature. We strongly prefer models that: 1) go back at least 20kyr, so that they can include the LGM; 2) include predictors from distinct ocean zones to allow multiple relevant climate signals to shape our reconstruction

# Records from all available Pacific regions that influence California precipitation. Multiple records exist for the Sea of Okhotsk, so we prefer the highest-ranked in loo AMONG those that also cover 20kyr, which is Davis5. Davis4 ranks higher overall but stops at 15kya, as does Davis8.
mld_SDPS <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int,
                data = clim,
                family = gaussian(),
                prior = paashi_priors,
                chains = 4, iter = 3000, warmup = 1000,
                seed = 2025)
summary(mld_SDPS)
saveRDS(mld_SDPS, "Models/mld_SDPS.rds")
#Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept     56.86      1.18    54.58    59.19 1.00     4887     4849
#Seki_Int       0.80      0.21     0.40     1.21 1.00     5262     5858
#Davis5_Int     1.22      0.22     0.79     1.65 1.00     6098     5719
#Pahnke_Int     1.21      0.42     0.37     2.04 1.00     4806     5217
#Stott_Int      0.22      0.12    -0.01     0.45 1.00     7161     5731

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     1.70      0.03     1.65     1.75 1.00     8371     5731
mld_SDPS <- readRDS("Models/mld_SDPS.rds")

#The previously submitted model, for continuity.
mld_Long <- brm(elev ~ Seki_Int + Stott_Int + Davis5_Int,
                data = clim,
                family = gaussian(),
                prior = paashi_priors,
                chains = 4, iter = 3000, warmup = 1000,
                seed = 2025)
summary(mld_Long)
saveRDS(mld_Long, "Models/mld_Long.rds")


##all 3-predictor geographically diverse subsets

#SPS: no Okhotsk. Okhotsk has the least-documented teleconnection to california precip
mld_SPS <- brm(elev ~ Seki_Int + Pahnke_Int + Stott_Int,
               data = clim,
               family = gaussian(),
               prior = paashi_priors,
               chains = 4, iter = 3000, warmup = 1000,
               seed = 2025)
summary(mld_SPS)
saveRDS(mld_SPS, "Models/mld_SPS.rds")

#DPS: no California Margin. not likely to perform well, losing most proximate record
mld_DPS <- brm(elev ~ Davis5_Int + Pahnke_Int + Stott_Int,
               data = clim,
               family = gaussian(),
               prior = paashi_priors,
               chains = 4, iter = 3000, warmup = 1000,
               seed = 2025)
summary(mld_DPS)
saveRDS(mld_DPS, "Models/mld_DPS.rds")

#SDP: no western tropical Pacific. western tropical pacific is relevant to el nino
mld_SDP <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int,
               data = clim,
               family = gaussian(),
               prior = paashi_priors,
               chains = 4, iter = 3000, warmup = 1000,
               seed = 2025)
summary(mld_SDP)
saveRDS(mld_SDP, "Models/mld_SDP.rds")

#Same 4-region structure as SDPS but swapping Davis5 for Davis3, the second-best performing okhotsk record. checks the importance of Davis5
mld_SD3PS <- brm(elev ~ Seki_Int + Davis3_Int + Pahnke_Int + Stott_Int,
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors,
                 chains = 4, iter = 3000, warmup = 1000,
                 seed = 2025)
summary(mld_SD3PS)
saveRDS(mld_SD3PS, "Models/mld_SD3PS.rds")

##Compare all combined models by LOO
loo(mld_Long, mld_SDPS,
    mld_SPS, mld_DPS, mld_SDP, mld_SD3PS, compare = TRUE)

#Model comparisons:
#          elpd_diff se_diff
#mld_SDPS    0.0       0.0  
#mld_SDP    -0.7       2.0  
#mld_SD3PS  -2.7       6.5  
#mld_Long   -3.3       2.8  
#mld_DPS    -6.8       3.9  
#mld_SPS   -13.8       5.4  

#uncertainty propagation is deferred to the final selected model. Once the best combined model is chosen, we fit it with ARMA and propagate chronological + interpolation uncertainty through the full hindcast.

```


### Figure 2: Study Area and Data Overview (assisted by Claude Code)
```{r fig2_map_and_sst, fig.width = 7.5, fig.height = 6}

##Let's visualize where our selected SSTs are coming from and what they look like.

# Panel A: Pacific Basin map with SST core sites + Tulare Lake
#the four predictors in the selected model, one per ocean region, plus the lake itself
sites <- data.frame(
  name = c("Pa'ashi", "Seki et al.\n(California Margin)",
           "Davis et al.\n(Sea of Okhotsk)",
           "Pahnke et al.\n(E. Tropical Pacific)",
           "Stott et al.\n(W. Equatorial Pacific)"),
  lon  = c(-119.5, -121.10, 148.32, -77.96, 133.44),
  lat  = c(35.9, 34.53, 50.40, 4.67, -5.00),
  type = c("lake", "sst", "sst", "sst", "sst"),
  col  = c(col_observed, col_seki, col_davis5, col_pahnke, col_stott)
)

world2 <- map_data("world2")
sites$lon360 <- ifelse(sites$lon < 0, sites$lon + 360, sites$lon)

fig1a <- ggplot() +
  geom_polygon(data = world2, aes(x = long, y = lat, group = group),
               fill = "gray92", color = "gray70", linewidth = 0.15) +
  geom_point(data = sites[1,], aes(x = lon360, y = lat),
             color = sites$col[1], shape = 18, size = 3.5) +
  geom_point(data = sites[-1,], aes(x = lon360, y = lat),
             color = sites$col[-1], fill = NA,
             shape = 21, size = 4.5, stroke = 1.5) +
  geom_text_repel(data = sites, aes(x = lon360, y = lat, label = name),
                  size = 2.2, color = sites$col,
                  box.padding = 0.5, max.overlaps = 10,
                  fontface = "bold", lineheight = 0.85) +
  coord_cartesian(xlim = c(100, 300), ylim = c(-15, 58), expand = FALSE) +
  scale_x_continuous(breaks = c(120, 150, 180, 210, 240, 270),
                     labels = c("120\u00B0E", "150\u00B0E", "180\u00B0",
                                "150\u00B0W", "120\u00B0W", "90\u00B0W")) +
  labs(title = "A. Study area and paleoclimate data sources",
       x = NULL, y = NULL) +
  theme_paashi() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 6),
        panel.grid.major = element_line(color = "gray85", linewidth = 0.2))

# Panel B: SST predictor time series
#the real measurements for each predictor. These are the bare record columns, which hold a value only at sampled years. We plot these rather than the _Int columns because the interpolated values are centered on each record's mean, which would stack all four series on top of each other.
sst_raw <- clim %>%
  select(Year_B2005, Seki, Davis5, Pahnke, Stott) %>%
  pivot_longer(-Year_B2005, names_to = "source", values_to = "sst_raw") %>%
  filter(!is.na(sst_raw)) %>%
  mutate(source = recode(source,
    "Seki"   = "Seki (California Margin)",
    "Davis5" = "Davis5 (Sea of Okhotsk)",
    "Pahnke" = "Pahnke (E. Tropical Pacific)",
    "Stott"  = "Stott (W. Equatorial Pacific)"))

sst_colors <- c(
  "Seki (California Margin)"       = col_seki,
  "Davis5 (Sea of Okhotsk)"        = col_davis5,
  "Pahnke (E. Tropical Pacific)"   = col_pahnke,
  "Stott (W. Equatorial Pacific)"  = col_stott)

fig1b <- ggplot() +
  #ggplot draws the connecting line itself, straight from one measurement to the next
  geom_line(data = sst_raw,
            aes(x = Year_B2005 - BP_OFFSET, y = sst_raw, color = source),
            linewidth = 0.3, alpha = 0.7) +
  geom_point(data = sst_raw,
             aes(x = Year_B2005 - BP_OFFSET, y = sst_raw, color = source),
             size = 0.5, alpha = 0.7) +
  scale_color_manual(values = sst_colors) +
  scale_x_reverse(labels = comma_format(),
                  breaks = seq(0, 28000, by = 4000)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Sea Surface Temperature (\u00B0C)",
       title = "B. Pacific SST records used as predictors",
       subtitle = "Points are the original measurements; lines connect them",
       color = NULL) +
  theme_paashi() +
  theme(legend.position = "bottom")

fig1 <- fig1a / fig1b + plot_layout(heights = c(3, 3))
fig1
save_fig(fig1, "Figure_2_Data_Overview.tiff", width_mm = 190, height_mm = 170)
```

##Add Prior Year Influence
```{r refining_models}
##Now that we have selected our SST sets with the best correlation to lake elevation, we want to see if other factors play in. Because this is a hydrological system, it is likely that the prior year can influence the next year.

##first, let's check our "auto correlation function" or ACF, to see how correlated each year's value is to the prior year; this will help us design better models.

#fix data for acf and pacf
clim <- clim[order(clim$Year_B2005), ] #make double sure clim is read in order of year
y <- clim$elev
y_complete <- stats::na.omit(y) #acf and pacf will not work if there are NAs

# acf checks auto correlation to see how long a value can influence subsequent values. PACF (partial ACF) removes the overlap so you can see more clearly the direct influence of prior years.
acf(y_complete, main = "ACF of elev (NAs dropped)")
pacf(y_complete, main = "PACF of elev (NAs dropped)")

#looks like we are strongly correlated for 1 year prior in the PACF. So, in ACF, value is strongly correlated with prior values, slowly tapering until ~15 prior values.
#periodicity around ~40 years backs up oral traditions, but is not necessarily strong enough to warrant adding fourier terms or seasonal arma. We will investigate periodicity in a bit.

## now we can run our best elev ~ SST model, adding influence from prior years. There are two arguments for adding in auto-correlation to a model. ar(p=x) says that the current value is correlated with past VALUES of x prior years. ma(q=x) says that the current value is correlated with past SHOCKS, or large CHANGES in value from the expected/mean value, from x prior years. arma(p = , q = ) just combines both. We can determine our p and q values from our ACF and PACF results.

#because we have a big spike in correlation for the previous year, plus a small correlation with the prior year in PACF; and a smooth taper in ACF, we can see that what matters most strongly is the VALUE of the prior year. If there were multiple spikes before taper in ACF, then we might pay more attention to our q value for ma().

#given what we see here, we can probably be safe with ar(p = 1), but could also reasonably try arma(p = 1, q = 1). 

##model SDPS with previous year effects
#ar(p = 1) for long, smooth persistence, good in hydrological models
mld_SDPSar <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int + ar(p = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = c(paashi_priors,
                           prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1)),
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_SDPSar)
#plot(mld_SDPSar)
#pp_check(mld_SDPSar, ndraws = 100)


#ma(q = 1) does the deviation from the mean matter more than the actual value? Let's check ma
mld_SDPSma <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int + ma(q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = c(paashi_priors,
                           prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1)),
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_SDPSma)
#plot(mld_SDPSma)
#pp_check(mld_SDPSma, ndraws = 100)

#arma(p = 1, q = 1) combines to see if both matter. This is our selected model.
mld_SDPSarma <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int +
                      arma(p = 1, q = 1, time = Year_B2005),
                    data = clim,
                    family = gaussian(),
                    prior = paashi_priors_arma,
                    chains = 4,
                    iter = 4000,
                    warmup = 2000,
                    seed = 2025,
                    control = list(adapt_delta = 0.95))

summary(mld_SDPSarma)
plot(mld_SDPSarma)
pp_check(mld_SDPSarma, ndraws = 100)
saveRDS(mld_SDPSarma, file = "Models/mld_SDPSarma.rds")
mld_SDPSarma <- readRDS("Models/mld_SDPSarma.rds")

#Correlation Structures:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#ar[1]     0.84      0.02     0.80     0.87 1.00     6326     5671
#ma[1]    -0.11      0.03    -0.16    -0.05 1.00     6652     5547

#Regression Coefficients:
#           Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept     57.75      3.71    50.47    65.05 1.00     7053     5111
#Seki_Int       0.68      0.66    -0.62     1.97 1.00     7431     5415
#Davis5_Int     1.21      0.73    -0.23     2.67 1.00     8161     5473
#Pahnke_Int     0.94      1.31    -1.65     3.50 1.00     6993     5635
#Stott_Int      0.10      0.38    -0.64     0.85 1.00     8150     5763

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     1.03      0.02     1.00     1.06 1.00     8785     5580

#compare no-autocorrelation, ar only, ma only, and both
loo(mld_SDPS, mld_SDPSar, mld_SDPSma, mld_SDPSarma, compare = TRUE)

#Model comparisons:
#             elpd_diff se_diff
#mld_SDPSarma    0.0       0.0 
#mld_SDPSar     -4.4       3.5 
#mld_SDPSma   -406.2      31.0 
#mld_SDPS     -999.3      42.6 

```

### Model Diagnostics Figure (assisted by Claude Code)
```{r figS3_diagnostics, fig.width = 7.5, fig.height = 8}
mld_SDPSarma <- readRDS("Models/mld_SDPSarma.rds")

# Panel A: Trace plots
s4a <- mcmc_trace(mld_SDPSarma,
                  pars = c("b_Intercept", "b_Seki_Int", "b_Davis5_Int",
                           "b_Pahnke_Int", "b_Stott_Int", "sigma", "ar[1]", "ma[1]"),
                  facet_args = list(ncol = 1)) +
  labs(title = "A. Trace plots", x = "Iteration", y = "Parameter value") +
  theme_paashi(base_size = 7)

# Panel B: Posterior predictive density overlay
s4b <- pp_check(mld_SDPSarma, ndraws = 100, type = "dens_overlay") +
  labs(title = "B. Posterior predictive check",
       x = "Lake Elevation (masl)", y = "Density") +
  theme_paashi(base_size = 7)

# Panel C: ACF posterior predictive check
yrep   <- posterior_predict(mld_SDPSarma, ndraws = 200)
y_obs  <- mld_SDPSarma$data$elev
acf_obs_vals <- acf(y_obs[!is.na(y_obs)], lag.max = 20, plot = FALSE)$acf[,,1]

acf_reps <- apply(yrep[1:200, !is.na(y_obs)], 1, function(y) {
  acf(y, lag.max = 20, plot = FALSE)$acf[,,1]
})

acf_rep_df <- as.data.frame(t(acf_reps)) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "lag", values_to = "acf") %>%
  mutate(lag = as.integer(gsub("V", "", lag)) - 1)

acf_obs_df <- data.frame(lag = 0:20, acf = acf_obs_vals)

s4c <- ggplot() +
  geom_line(data = acf_rep_df, aes(x = lag, y = acf, group = draw),
            alpha = 0.08, color = "gray50") +
  geom_line(data = acf_obs_df, aes(x = lag, y = acf),
            color = "black", linewidth = 0.8) +
  labs(x = "Lag (years)", y = "Autocorrelation",
       title = "C. Posterior predictive ACF check",
       subtitle = "Black = observed; gray = simulated") +
  theme_paashi(base_size = 7)

# Panel D: LOO Pareto k diagnostics
loo_sdpsarma <- loo(mld_SDPSarma)
k_values <- loo_sdpsarma$diagnostics$pareto_k
k_df <- data.frame(obs = seq_along(k_values), k = k_values)

s4d <- ggplot(k_df, aes(x = obs, y = k)) +
  geom_point(size = 0.3, alpha = 0.4, color = col_model_inner) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange",
             linewidth = 0.4) +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "red",
             linewidth = 0.4) +
  annotate("text", x = nrow(k_df) * 0.85, y = 0.55,
           label = "k = 0.5 (caution)", size = 2.2, color = "orange") +
  annotate("text", x = nrow(k_df) * 0.85, y = 0.75,
           label = "k = 0.7 (unreliable)", size = 2.2, color = "red") +
  labs(x = "Observation", y = "Pareto k",
       title = "D. LOO-CV Pareto k diagnostics",
       subtitle = paste0("All ", nrow(k_df), " observations below 0.5")) +
  theme_paashi(base_size = 7)

figS4 <- (s4a | (s4b / s4c / s4d)) +
  plot_layout(widths = c(1, 1)) +
  plot_annotation(
    title = "Supplementary Figure S3: Model Diagnostics (mld_SDPSarma)")
figS4
save_fig(figS4, "Figure_S3_Diagnostics.tiff", width_mm = 190, height_mm = 220)
```

## Continue Model Testing
```{r model_testing_continued}
# Test whether SSTs are necessary, or if ARMA alone suffices
# No "b" class here since there are no slope predictors (intercept-only)
mld_arma_only <- brm(elev ~ 1 + arma(p = 1, q = 1, time = Year_B2005),
                     data = clim,
                     family = gaussian(),
                     prior = c(
                       prior(normal(63, 5), class = "Intercept"),
                       prior(student_t(3, 0, 3), class = "sigma"),
                       prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1),
                       prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1)),
                     chains = 4, iter = 4000, warmup = 2000,
                     seed = 2025,
                     control = list(adapt_delta = 0.95))

loo(mld_SDPSarma, mld_arma_only, compare = TRUE)
#essentially identical for year-to-year prediction. This is expected in loo since ARMA dominates year over year correlation; loo (leave one out) observes prediction when one observation is left out at a time. SSTs are essential for century+ scale climate forcing, because arma alone does not inform the mean when making predictions on new data.

##let's check for non-linear and more complex relationships

#spline instead of linear
mld_SDPSarmas <- brm(elev ~ s(Seki_Int) + s(Davis5_Int) + s(Pahnke_Int) + s(Stott_Int) + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_SDPSarmas)
#plot(mld_SDPSarmas)
#pp_check(mld_SDPSarmas, ndraws = 100)
saveRDS(mld_SDPSarmas, file = "Models/mld_SDPSarmas.rds")

#Smoothing Spline Hyperparameters:
#                   Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sds(sSeki_Int_1)       2.02      1.82     0.07     6.68 1.00     2517     4699
#sds(sDavis5_Int_1)     1.30      1.11     0.04     4.07 1.00     3460     3590
#sds(sPahnke_Int_1)     1.45      1.21     0.05     4.54 1.00     3568     3163
#sds(sStott_Int_1)      1.06      0.94     0.03     3.39 1.00     3881     3527

#Correlation Structures:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#ar[1]     0.83      0.02     0.79     0.86 1.00     5453     5669
#ma[1]    -0.10      0.03    -0.16    -0.04 1.00     7361     5965

#Regression Coefficients:
#              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept        62.21      0.12    61.97    62.45 1.00    10415     5703
#sSeki_Int_1       2.00      3.13    -3.83     8.89 1.00     5408     4652
#sDavis5_Int_1     0.08      2.80    -6.42     4.82 1.00     5519     5582
#sPahnke_Int_1    -1.19      3.18    -8.28     4.25 1.00     4719     5280
#sStott_Int_1     -0.54      2.55    -6.52     3.91 1.00     5060     4953

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     1.03      0.02     1.00     1.06 1.00    10956     6223

#linear interactions between variables
# Using * for full interaction hierarchy (main effects + 2-way + 3-way)
mld_SDPSarmar <- brm(elev ~ Seki_Int * Davis5_Int * Pahnke_Int * Stott_Int + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_SDPSarmar)
#plot(mld_SDPSarmar)
#pp_check(mld_SDPSarmar, ndraws = 100)
saveRDS(mld_SDPSarmar, file = "Models/mld_SDPSarmar.rds")

#Correlation Structures:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#ar[1]     0.83      0.02     0.80     0.87 1.00     6732     5806
#ma[1]    -0.10      0.03    -0.16    -0.04 1.00     8246     6198

#Regression Coefficients:
#                                         Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept                                   60.12      9.54    41.54    78.83 1.00     7753     5627
#Seki_Int                                     0.78      3.17    -5.50     6.99 1.00     4958     4874
#Davis5_Int                                  -2.06      4.75   -11.29     7.09 1.00    10276     6136
#Pahnke_Int                                   0.12      3.88    -7.51     7.69 1.00     6897     5969
#Stott_Int                                    1.44      4.41    -7.21     9.97 1.00     8393     6036
#Seki_Int:Davis5_Int                         -2.72      2.96    -8.59     3.08 1.00     4546     5479
#Seki_Int:Pahnke_Int                          0.01      2.10    -4.07     4.14 1.00     4400     4980
#Davis5_Int:Pahnke_Int                       -1.28      3.91    -9.02     6.50 1.00     7179     5981
#Seki_Int:Stott_Int                           1.89      2.53    -3.07     6.93 1.00     4313     4978
#Davis5_Int:Stott_Int                        -0.85      4.63    -9.94     8.01 1.00    10322     6149
#Pahnke_Int:Stott_Int                         2.39      3.18    -3.79     8.61 1.00     5962     5648
#Seki_Int:Davis5_Int:Pahnke_Int               2.42      2.34    -2.16     6.93 1.00     4602     5179
#Seki_Int:Davis5_Int:Stott_Int                0.57      3.07    -5.46     6.52 1.00     4891     5162
#Seki_Int:Pahnke_Int:Stott_Int               -2.70      2.10    -6.85     1.44 1.00     4559     4968
#Davis5_Int:Pahnke_Int:Stott_Int             -0.67      3.90    -8.34     6.94 1.00     6582     6394
#Seki_Int:Davis5_Int:Pahnke_Int:Stott_Int     0.64      2.29    -3.91     5.16 1.00     4788     5685

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     1.03      0.02     1.00     1.06 1.00    12185     5750

loo(mld_SDPSarma, mld_SDPSarmas, mld_SDPSarmar, compare = TRUE)

#Model comparisons:
#              elpd_diff se_diff
#mld_SDPSarmas  0.0       0.0   
#mld_SDPSarma  -0.1       1.4   
#mld_SDPSarmar -1.2       1.5 

# Differences small; keep simpler mld_SDPSarma for ease of processing and interpreting

##let's see if variance and flood frequency are influenced by SST
#best linear model plus variance tested
#note the bf() part, this is used if there are multiple parts in the model
mld_SDPSarma_sig <- brm(bf(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int + arma(p = 1, q = 1, time = Year_B2005),
    # variance (sigma) model: does variability depend on SST?
    sigma ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int),
  data   = clim,
  family = gaussian(),
  prior = c(
    prior(normal(63, 5), class = "Intercept"),
    prior(normal(0, 5), class = "b"),
    prior(normal(0, 1), class = "b", dpar = "sigma"),
    prior(student_t(3, 0, 1), class = "Intercept", dpar = "sigma")),
  chains = 4,
  iter   = 5000,
  warmup = 2500,
  seed = 2025,
  control = list(adapt_delta = 0.95))

saveRDS(mld_SDPSarma_sig, file = "Models/mld_SDPSarma_sig.rds")
readRDS("Models/mld_SDPSarma_sig.rds") -> mld_SDPSarma_sig
summary(mld_SDPSarma_sig)
# remember that sigma summaries default to log.
##Bad fit; SSTs do not meaningfully influence year-to-year variance when ARMA is present.

# Skew-normal model: check if lake elevation distribution is asymmetric.
# Note: skew-normal + ARMA was too heavy
mld_SDPS_skew <- brm(bf(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int,
    # right tail (alpha) model: do high elevations depend on SST?
    alpha ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int),
  data   = clim,
  family = skew_normal(),
  chains = 4,
  iter   = 3000,
  warmup = 1000,
  seed = 2025)

saveRDS(mld_SDPS_skew, file = "Models/mld_SDPS_skew.rds")
readRDS("Models/mld_SDPS_skew.rds") -> mld_SDPS_skew
summary(mld_SDPS_skew)

#Regression Coefficients:
#                 Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#Intercept           56.70      1.16    54.43    58.98 1.00     7206     6176
#alpha_Intercept     22.39      8.62     6.15    40.86 1.00     6032     4026
#Seki_Int             0.83      0.21     0.43     1.24 1.00     7814     6653
#Davis5_Int           1.22      0.23     0.77     1.67 1.00     9026     5995
#Pahnke_Int           1.19      0.42     0.37     2.00 1.00     6981     6151
#Stott_Int            0.35      0.12     0.10     0.60 1.00     9834     5583
#alpha_Seki_Int      -2.85      1.42    -5.86    -0.19 1.00     6323     4056
#alpha_Davis5_Int    -4.88      1.63    -8.21    -1.82 1.00     7076     5473
#alpha_Pahnke_Int    -6.08      3.19   -12.98    -0.01 1.00     5722     4237
#alpha_Stott_Int     -2.34      0.80    -3.92    -0.78 1.00     7902     5612

#Further Distributional Parameters:
#      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
#sigma     1.71      0.03     1.66     1.77 1.00     9673     6136

## All four SSTs contribute credibly to the mean; all four alpha (skewness) terms exclude zero: Seki -2.85 [-5.86, -0.19], Davis5 -4.88 [-8.21, -1.82], Pahnke -6.08 [-12.98, -0.01], Stott -2.34 [-3.92, -0.78]. So warmer SSTs are associated with a less right-skewed distribution of lake states across ALL four regions. That is consistent with an ENSO teleconnection (higher temperatures, higher and more consistent lake levels), but the effect is not specific to one ocean region.
## NOTE: likely dropping the skew-normal discussion from the manuscript, as it adds little at this point.

```

## Uncertainty Propagation
```{r uncertainty_propagation}
##Now that we have finalized the model we want to use, we can finally propagate the uncertainty from the draws we created up top into our model. But, if we simply take the information from all the draws equally, the underlying signal can be totally washed out, or "attenuated," as discussed by McKay. One way around this is a "latent state space model," where the shape of random draws is constrained by how well they fit the outcome variable: essentially, our outcome variable gives us information about our predictors, constraining our uncertainty. Constraining draws this way takes a lot of custom STAN code, so instead, we will do a less efficient approach in R. We will fit the model to each draw of uncertainty, then rank them by how well they predict our outcome variable (log marginal likelihood). Then, we will weight the models according to their ranks, and pull posterior draws from them accordingly. We create mld_SDPSarma_prop, a list holding the weighted curves, weights, and everything else needed for predictions downstream.

#the four predictors in the selected model
SDPS <- c("Seki", "Davis5", "Pahnke", "Stott")

#center each record's draws on the same constant used for the _Int columns
draws_centered <- lapply(SDPS, function(v) interp[[v]]$draws - sst_means[v])
names(draws_centered) <- SDPS

#prep the data for the model
draw_data <- function(i) {
  d <- data.frame(elev = clim$elev, Year_B2005 = clim$Year_B2005)
  for (v in SDPS) d[[paste0(v, "_Int")]] <- draws_centered[[v]][, i]
  d
}

#run all chains in parallel to save time
options(mc.cores = parallel::detectCores())

#the fits and their bridge-sampling scores each take a while, so both are saved and reloaded if available
dir.create("Models", showWarnings = FALSE)
fits_file  <- sprintf("Models/prop_fits_n%d_meas%s_age%s.rds",
                      N_DRAWS, MEAS_SD, AGE_SD)
logml_file <- sprintf("Models/prop_logml_n%d_meas%s_age%s.rds",
                      N_DRAWS, MEAS_SD, AGE_SD)

if (file.exists(fits_file)) {
  message("reusing ", fits_file)
  fits <- readRDS(fits_file)
} else {
  
  ##fit draw 1 to compile the model. save_pars(all = TRUE) is required for bridge_sampler() later.
  fit1 <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int +
                arma(p = 1, q = 1, time = Year_B2005),
              data = draw_data(1),
              family = gaussian(),
              prior = paashi_priors_arma,
              chains = 4, iter = 4000, warmup = 2000,
              seed = 2025, refresh = 0,
              control = list(adapt_delta = 0.95),
              save_pars = save_pars(all = TRUE))

  ## fit remaining draws using the compiled model
  fits <- vector("list", N_DRAWS)
  fits[[1]] <- fit1
  for (i in 2:N_DRAWS) {
    message("fitting draw ", i, " of ", N_DRAWS)
    fits[[i]] <- update(fit1, newdata = draw_data(i),
                        recompile = FALSE, refresh = 0)
  }
  #xz compression at level 9 keeps this file under GitHub's 100 MB limit (compress = "xz" alone uses level 6, which is not enough); readRDS() reads it the same way
  con <- xzfile(fits_file, "wb", compression = 9)
  saveRDS(fits, con)
  close(con)
}

#score each fit with bridge_sampler to get log marginal likelihood
if (file.exists(logml_file)) {
  message("reusing ", logml_file)
  logml <- readRDS(logml_file)
} else {
  logml <- numeric(N_DRAWS)
  for (i in seq_len(N_DRAWS)) {
    message("bridge sampling draw ", i, " of ", N_DRAWS)
    logml[i] <- as.numeric(bridge_sampler(fits[[i]], silent = TRUE)$logml)
  }
  saveRDS(logml, logml_file)
}

#turn scores into weights. subtract the max/best mean likelihood to make the numbers usable, then get the exponent to each likelihood as a fraction of the best one, then divide by the total to make them all add up to 1.
w <- exp(logml - max(logml))
w <- w / sum(w)

#how concentrated are the weights? sort the draws heaviest to lightest and add them up. n_to_one is how many it takes for the total draws to reach weight 1
w_ord    <- order(w, decreasing = TRUE)
w_sorted <- w[w_ord]
w_cum    <- cumsum(w_sorted)
#WT_DIGITS decides how close to 1 counts as 1. (right now it's .99)
WT_DIGITS <- 2
n_to_one <- which(round(w_cum, WT_DIGITS) >= 1)[1]

cat("draws:", N_DRAWS,
    " heaviest weight:", round(max(w), 3),
    " draws to a weight of 1:", n_to_one, "\n\n")

#Draws by weight
print(head(data.frame(draw       = w_ord,
                      weight     = round(w_sorted, 4),
                      cumulative = round(w_cum, 4)), n_to_one),
      row.names = FALSE)

#extract posterior mean coefficients from each fit
coefs <- sapply(fits, function(f) fixef(f)[, "Estimate"])


## For each fit, calculate the mean SST line by getting the average intercept and slopes and calculating an elev value for each year. Similar to fitted() but faster.
curve_one <- function(i) {
  b <- coefs[, i]
  X <- as.matrix(draw_data(i)[, paste0(SDPS, "_Int")])
  as.vector(b["Intercept"] + X %*% b[paste0(SDPS, "_Int")])
}
curves <- sapply(seq_len(N_DRAWS), curve_one)

#apply the weights we just calculated to each fit, so the worst ones end up close to 0. They all add up to 1, so the result is a weighted average of all the mean lines according to their value.
mean_w <- as.vector(curves %*% w)

#The curves above use only the posterior MEAN, so they do not show uncertainty. To show uncertainty, we pool posterior draws from all of our fits together, but, we pool draws according to that fit's assigned weight.

N_POOL <- 5000   #total posterior draws in the pool. Each fit has 8,000 draws so don't go above
#determine how many draws each fit supplies from its weight
n_take <- round(w * N_POOL)

#pick which posterior draws to keep from each fit, and pull the noise parameters; keep the row numbers (kept_draws) so the linear predictor and the noise always come from the same posterior draw.
set.seed(2025)
kept_draws <- vector("list", N_DRAWS)
post_noise <- vector("list", N_DRAWS)
for (i in seq_len(N_DRAWS)) {
  d <- as.data.frame(fits[[i]])
  kept_draws[[i]] <- sample(nrow(d), n_take[i])
  post_noise[[i]] <- as.matrix(d[kept_draws[[i]], c("sigma", "ar[1]", "ma[1]")])
}

## predict_weighted gives the SST-driven prediction for a chosen set of years (rows). It draws one curve for every posterior draw in the pool, then reads the 2.5% and 97.5% quantiles for each year.
predict_weighted <- function(rows) {

  n_years <- length(rows)
  n_curve <- sum(n_take)

  #one column per posterior draw in the pool
  curves <- matrix(NA_real_, nrow = n_years, ncol = n_curve)

  ##build the curves
  col <- 0
  for (i in seq_len(N_DRAWS)) {
    #skip fits that supply no draws to the pool
    if (n_take[i] == 0) next
    #the SST-driven mean for each kept posterior draw. incl_autocor = FALSE leaves the ARMA term out because it adds nothing to the mean. draw_ids pulls the relevant info from post_noise.
    mu <- posterior_linpred(fits[[i]],
                            newdata      = draw_data(i)[rows, ],
                            incl_autocor = FALSE,
                            draw_ids     = kept_draws[[i]])
    cols <- col + seq_len(n_take[i])
    curves[, cols] <- t(mu)
    col <- col + n_take[i]
  }

  ##the 95% CI for the mean
  fit_lo <- numeric(n_years)
  fit_hi <- numeric(n_years)
  for (t in seq_len(n_years)) {
    q <- quantile(curves[t, ], c(0.025, 0.975))
    fit_lo[t] <- q[1]
    fit_hi[t] <- q[2]
  }

  #add an ARMA(1,1) path simulated from that curve's own ar, ma and sigma; add the noise into `curves`
  col <- 0
  for (i in seq_len(N_DRAWS)) {
    for (j in seq_len(n_take[i])) {
      col <- col + 1
      b <- post_noise[[i]][j, ]
      noise <- arima.sim(list(ar = b["ar[1]"], ma = b["ma[1]"]),
                         n = n_years, sd = b["sigma"])
      curves[, col] <- curves[, col] + as.numeric(noise)
    }
  }

  ##the 95% prediction interval
  pred_lo <- numeric(n_years)
  pred_hi <- numeric(n_years)
  for (t in seq_len(n_years)) {
    q <- quantile(curves[t, ], c(0.025, 0.975))
    pred_lo[t] <- q[1]
    pred_hi[t] <- q[2]
  }

  fit_q  <- cbind(fit_lo, fit_hi)
  pred_q <- cbind(pred_lo, pred_hi)
  data.frame(Fit_Est = mean_w[rows],
             Fit_SE = (fit_q[, 2] - fit_q[, 1]) / (2 * 1.96),
             Fit_Q2.5 = fit_q[, 1],
             Fit_Q97.5 = fit_q[, 2],
             Pred_Est = mean_w[rows],
             Pred_SE = (pred_q[, 2] - pred_q[, 1]) / (2 * 1.96),
             Pred_Q2.5 = pred_q[, 1],
             Pred_Q97.5 = pred_q[, 2])
}

## bundle everything into a list for downstream predictions
mld_SDPSarma_prop <- list(
  curves = curves,
  w = w,
  logml = logml,
  coefs = coefs,
  mean_w = mean_w,
  kept_draws = kept_draws,
  post_noise = post_noise,
  n_draws = N_DRAWS,
  n_take = n_take)

```

## Blind Prediction of Calibration Period
```{r blind_training_pred}
## Predict the calibration period using the weighted pool. This is comparable to the hindcast: ARMA is not used in the mean, so the predicted mean is purely SST-driven, and ARMA noise is simulated for the prediction interval. We can compare this to observed elevations to see how well the weighted model recovers the calibration signal.

#load the base model
mld_SDPSarma <- readRDS("Models/mld_SDPSarma.rds")

#rows where we have observed lake elevations
cal_rows <- which(!is.na(clim$elev))

#weighted mean and 95% prediction interval for the calibration years
set.seed(2025)
cal_pred <- predict_weighted(cal_rows)

##build the blind prediction data frame
blind_preds <- data.frame(
  year_bp  = clim$Year_B2005[cal_rows] - BP_OFFSET,   #cal yr BP for plotting
  elev_obs = clim$elev[cal_rows],
  Estimate = cal_pred$Pred_Est,
  Q2.5     = cal_pred$Pred_Q2.5,
  Q97.5    = cal_pred$Pred_Q97.5
)

fig_blind_train <- ggplot(blind_preds, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.4, alpha = 0.4) +
  geom_smooth(aes(y = elev_obs), method = "loess", span = 400 / nrow(blind_preds),
              color = col_observed, linewidth = 0.9, se = FALSE) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.8) +
  scale_x_reverse(labels = comma_format()) +
  labs(x = "Years Before Present (cal yr BP)", y = "Lake Elevation (masl)",
       title = "Full Model: Blind Prediction of Calibration Period",
       subtitle = "Weighted pool; SST-driven mean only; ARMA excluded from the mean, simulated for the interval") +
  theme_paashi()
fig_blind_train
save_fig(fig_blind_train, "Figure_3_Blind_Calibration.tiff")

```

## Exploring Periodicity- Persistence, Climate Forcing, Basin Physics; What do our CIs contain?
```{r exploring_periodicity}
## There is still a lot of noise in lake elevations (represented in our CIs), even when predicting from our current SST sets and with ARMA. Arma helps with understanding persistence of lake levels, but does not EXPLAIN deviations from the mean. When we actually look at a plot of lake elevation and listen to oral traditions, we know that the lake fluctuates year to year and has decadal and "generational" patterns. We should examine these patterns to see if we can explain shorter term variation / noise before we model deep into the past.

##Visual check: plot observed lake elevations with a line representing flood level; note ~decadal fluctuations and "generational" flood events.
clim <- clim[order(clim$Year_B2005), ] #make double sure clim is read in order of year

drainage_level <- 64.5
obs_elev <- clim %>% filter(!is.na(elev))

fig_drainage <- ggplot(obs_elev, aes(x = Calendar_Year)) +
  geom_line(aes(y = elev), color = "gray70", linewidth = 0.3) +
  geom_line(aes(y = pmax(elev, drainage_level)),
            color = col_observed, linewidth = 0.4) +
  #geom_hline(yintercept = drainage_level, color = col_observed,
             #linewidth = 0.3, linetype = "solid") +
  labs(x = "Calendar Year AD", y = "Water Surface Elevation (masl)",
       title = "Observed Pa'ashi Elevation — Drainage Periods Highlighted") +
  theme_paashi()
fig_drainage
save_fig(fig_drainage, "Figure_6_Drainage.tiff", width_mm = 190, height_mm = 80)

# Smoothed periodogram of observed elevations
y <- clim$elev
y_complete <- stats::na.omit(y)

sp <- spec.pgram(y_complete, spans = c(7, 7), plot = FALSE)
sp_period <- 1 / sp$freq
sp_mask <- sp_period >= 10 & sp_period <= 200
plot(sp_period[sp_mask], sp$spec[sp_mask], type = "l",
     xlab = "Period (years)", ylab = "Spectrum",
     main = "Smoothed Periodogram (10–200 year periods)")
##Two clear peaks at ~35-40 years and ~70-80 years. The 70-80 year peak is likely the harmonic (double) of the ~40-year cycle. Rise at 150+ is century-scale variance, probably SST-driven.

##Now let's assess the flood data quantitatively:

# Flood recurrence: how often does elev exceed 64.5m? min_gap_years: floods separated by fewer than this many years are grouped as one event to account for 1-2 year dips during a wet period
flood_threshold <- 64.5
min_gap_years <- 5
flood_years <- which(y_complete > flood_threshold)
flood_starts <- flood_years[c(TRUE, diff(flood_years) > min_gap_years)]
flood_gaps <- diff(flood_starts)
cat("Number of flood events (min gap =", min_gap_years, "years):", length(flood_starts), "\n")
cat("Mean return interval:", round(mean(flood_gaps), 1), "years\n")
cat("Median return interval:", round(median(flood_gaps), 1), "years\n")
cat("Range:", min(flood_gaps), "-", max(flood_gaps), "years\n")
flood_gap_df <- data.frame(gap = flood_gaps)
flood_mean <- mean(flood_gaps)
flood_median <- median(flood_gaps)
n_floods <- length(flood_starts)

fig_flood <- ggplot(flood_gap_df, aes(x = gap)) +
  geom_histogram(bins = 15, fill = col_model_ribbon, color = "white") +
  geom_vline(xintercept = flood_mean, linetype = "dashed", color = col_model_line, linewidth = 0.5) +
  geom_vline(xintercept = flood_median, linetype = "dotted", color = col_arch, linewidth = 0.5) +
  annotate("text", x = flood_mean + 2, y = Inf, vjust = 2, hjust = 0, size = 2.5,
           label = paste0("Mean = ", round(flood_mean, 1), " yr"), color = col_model_line) +
  annotate("text", x = flood_median - 2, y = Inf, vjust = 3.5, hjust = 1, size = 2.5,
           label = paste0("Median = ", round(flood_median, 1), " yr"), color = col_arch) +
  annotate("text", x = Inf, y = Inf, vjust = 1.5, hjust = 1.1, size = 2.5, color = "gray30",
           label = paste0("n = ", n_floods, " events over ~2,000 yr\n",
                          "Range: ", min(flood_gaps), "-", max(flood_gaps), " yr")) +
  labs(x = "Years Between Flood Events", y = "Count",
       title = paste0("Flood Return Intervals (elev > ", flood_threshold, " masl)")) +
  theme_paashi()
fig_flood
save_fig(fig_flood, "Figure_S5_Flood_Return.tiff", height_mm = 90)
# Result: 51 flood events over ~2,000 years. Mean return = 39.4 years, median = 32 years. Mode around 20-30 years. Right-skewed: most floods are generational (~25-40 years apart) but some gaps stretch to 80-120 years. Consistent with oral traditions of "once a generation" flooding. The right skew may reflect prolonged drought periods.

```

## PDO/ENSO Diagnostic: What Do Our Hindcast CIs Represent?
```{r pdo_enso_test}
## Our century-scale SSTs force long-term mean lake levels. ARMA(1,1) accounts for year-to-year persistence because the basin buffers swings in rainfall. But there are still year-over-year fluctuations and ~35-40 year periodicity that our models do not explain.

# Based on contemporary climate science, PDO (Pacific Decadal Oscillation) and ENSO (El Nino-Southern Oscillation) likely drive those fluctuations. They are the annually resolved versions of the same Pacific climate signals our century-scale proxies show:
# Stott_Int (~50-100yr resolution) = long-term mean of western equatorial Pacific, where ENSO operates; ~7 year cycles
# Seki_Int (~260yr resolution) = long-term mean of NE Pacific / California margin, where PDO operates; ~20-30 year cycles

# To quantify what our hindcast CIs represent, we fit three models on the ~1,000-year window where annually-resolved PDO and ENSO reconstructions overlap with our calibration data, then compare blind predictions (elev = NA, same as hindcast).
# Sources: MacDonald & Case 2005 (PDO, 993-1996 CE), Li et al. 2011 (ENSO, 900-2002 CE)

# Filter to rows where all predictors + PDO + ENSO are available (~1,000 years)
clim %>%
  filter(!is.na(elev), !is.na(Seki_Int), !is.na(Stott_Int),
         !is.na(Davis5_Int), !is.na(Pahnke_Int),
         !is.na(PDO), !is.na(ENSO)) %>%
  arrange(Year_B2005) -> clim_test

# Model 1: SSTs + ARMA baseline (same structure as mld_SDPSarma)
if (file.exists("Models/mld_test_sdpsarma.rds")) {
  mld_test_sdpsarma <- readRDS("Models/mld_test_sdpsarma.rds")
} else {
  mld_test_sdpsarma <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int + arma(p = 1, q = 1, time = Year_B2005),
      data = clim_test, family = gaussian(),
      prior = paashi_priors_arma,
      chains = 4, iter = 4000, warmup = 2000,
      seed = 2025,
      control = list(adapt_delta = 0.95))
  saveRDS(mld_test_sdpsarma, "Models/mld_test_sdpsarma.rds")
}

# Model 2: PDO + ENSO only (no SSTs). can annually-resolved indices alone track the lake?
if (file.exists("Models/mld_test_pdo_enso_only.rds")) {
  mld_test_pdo_enso_only <- readRDS("Models/mld_test_pdo_enso_only.rds")
} else {
  mld_test_pdo_enso_only <- brm(elev ~ PDO + ENSO + arma(p = 1, q = 1, time = Year_B2005),
      data = clim_test, family = gaussian(),
      prior = paashi_priors_arma,
      chains = 4, iter = 4000, warmup = 2000,
      seed = 2025,
      control = list(adapt_delta = 0.95))
  saveRDS(mld_test_pdo_enso_only, "Models/mld_test_pdo_enso_only.rds")
}

# Model 3: SSTs + PDO + ENSO + ARMA (full model for LOO reference)
if (file.exists("Models/mld_test_pdo_enso.rds")) {
  mld_test_pdo_enso <- readRDS("Models/mld_test_pdo_enso.rds")
} else {
  mld_test_pdo_enso <- brm(elev ~ Seki_Int + Davis5_Int + Pahnke_Int + Stott_Int + PDO + ENSO + arma(p = 1, q = 1, time = Year_B2005),
      data = clim_test, family = gaussian(),
      prior = paashi_priors_arma,
      chains = 4, iter = 4000, warmup = 2000,
      seed = 2025,
      control = list(adapt_delta = 0.95))
  saveRDS(mld_test_pdo_enso, "Models/mld_test_pdo_enso.rds")
}

# LOO comparison
loo(mld_test_sdpsarma, mld_test_pdo_enso_only, mld_test_pdo_enso)

# Coefficient comparison
summary(mld_test_sdpsarma)
summary(mld_test_pdo_enso_only)
summary(mld_test_pdo_enso)
## Loo and summaries show that SSTs don't add anything to the PDO/ENSO predictions. This is to be expected, as the SSTs are essentially just a less detailed record of the same climate signals.

# Blind prediction comparison: set elev = NA so ARMA starts from zero, same as the hindcast.
clim_test_blind <- clim_test %>% mutate(elev_obs = elev, elev = NA)

## This whole section uses the base mean fits for simplicity; any annual signal from the SSTs would be an artifact of our random interpolation process
blind_sst <- as.data.frame(predict(mld_test_sdpsarma, newdata = clim_test_blind))
blind_sst$year_bp  <- clim_test_blind$Year_B2005 - BP_OFFSET   #cal yr BP for plotting
blind_sst$elev_obs <- clim_test_blind$elev_obs

blind_pdo <- as.data.frame(predict(mld_test_pdo_enso_only, newdata = clim_test_blind))
blind_pdo$year_bp  <- clim_test_blind$Year_B2005 - BP_OFFSET   #cal yr BP for plotting
blind_pdo$elev_obs <- clim_test_blind$elev_obs

## SST+ARMA vs PDO+ENSO+ARMA
p_blind_sst <- ggplot(blind_sst, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3, alpha = 0.4) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.5) +
  scale_x_reverse(labels = comma_format()) +
  labs(x = "Years Before Present (cal yr BP)", y = "Lake Elevation (masl)",
       title = "Blind Prediction: SSTs + ARMA(1,1)") +
  theme_paashi()

p_blind_pdo <- ggplot(blind_pdo, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3, alpha = 0.4) +
  geom_line(aes(y = Estimate), color = col_pdo_line, linewidth = 0.5) +
  scale_x_reverse(labels = comma_format()) +
  labs(x = "Years Before Present (cal yr BP)", y = "Lake Elevation (masl)",
       title = "Blind Prediction: PDO + ENSO Only (No SSTs)") +
  theme_paashi()

p_blind_sst / p_blind_pdo
save_fig(p_blind_sst / p_blind_pdo, "Figure_S7_Blind_SST_vs_PDO.tiff", height_mm = 140)

# SST+ARMA follows multi-century mean; PDO+ENSO tracks decadal swings in the predicted mean. The structure visible in the PDO+ENSO panel is what our SST-based hindcast CIs represent.

## PDO is strongly related to lake elevation (+0.79m per unit [0.68, 0.91]), ENSO adds +0.20m per unit [0.16, 0.25]. LOO confirms PDO+ENSO improves fit substantially over SSTs alone (~122 elpd). Blind predictions show that PDO+ENSO produces year-to-year and decadal variation in the predicted mean, because it is annually-resolved Pacific climate data rather than a multi-century average.

# NOTE, these PDO and ENSO records only cover ~1,000 years, so they cannot be used for the 20,000-year hindcast. The year-to-year and decadal variability they explain is what our hindcast prediction intervals represent.

# Residual periodograms: does the ~40yr peak survive after removing climate forcing?
# Panel A: SSTs + ARMA(1,1) residuals (from mld_SDPSarma, full calibration period)
arma_resids <- residuals(mld_SDPSarma, type = "ordinary")[, "Estimate"]
sp_resid <- spec.pgram(arma_resids, spans = c(7, 7), plot = FALSE)
sp_resid_period <- 1 / sp_resid$freq
sp_resid_mask <- sp_resid_period >= 10 & sp_resid_period <= 200
df_sp_sst <- data.frame(period = sp_resid_period[sp_resid_mask],
                        spectrum = sp_resid$spec[sp_resid_mask])

fig_periodo_sst <- ggplot(df_sp_sst, aes(x = period, y = spectrum)) +
  geom_line(color = col_model_line, linewidth = 0.5) +
  labs(x = "Period (years)", y = "Spectrum",
       title = "A. Residual Periodogram: SSTs + ARMA(1,1)") +
  theme_paashi()

# Panel B: PDO + ENSO + ARMA(1,1) residuals (~1,000yr subset)
pdo_resids <- residuals(mld_test_pdo_enso_only, type = "ordinary")[, "Estimate"]
sp_pdo <- spec.pgram(pdo_resids, spans = c(7, 7), plot = FALSE)
sp_pdo_period <- 1 / sp_pdo$freq
sp_pdo_mask <- sp_pdo_period >= 10 & sp_pdo_period <= 200

df_sp_pdo <- data.frame(period = sp_pdo_period[sp_pdo_mask],
                        spectrum = sp_pdo$spec[sp_pdo_mask])

fig_periodo_pdo <- ggplot(df_sp_pdo, aes(x = period, y = spectrum)) +
  geom_line(color = col_pdo_line, linewidth = 0.5) +
  labs(x = "Period (years)", y = "Spectrum",
       title = "B. Residual Periodogram: PDO + ENSO + ARMA(1,1)") +
  theme_paashi()

fig_periodograms <- fig_periodo_sst / fig_periodo_pdo
fig_periodograms
save_fig(fig_periodograms, "Figure_S6_Residual_Periodograms.tiff", height_mm = 140)

```

## Physical Origin of Periodicity- Examine Adams (2015) Hydrological Model
```{r adams_physics_sims}
#So, the periodicity that survives in our residuals after removing SSTs + ARMA(1,1) is ALSO not a climate signal, as it remains even after using PDO and ENSO. We can test whether it is a physical property of the lake basin. Adams (2015) modeled Tulare Lake as a terminal basin where water only leaves via evaporation. He also notes ~32-64 year periodicity, and suggests that it is a combination of climate signals filtered through the lake's unique hydrology. 

#This idea makes  a lot of sense because surface area scales nonlinearly with volume, so linear increases in water volume or elevation create exponential increases in surface area, massively increasing evaporation. large lake ->  high evaporation -> lake shrinks -> less surface area -> less evaporation -> lake grows. This feedback likely produces the ~32-64 year oscillations regardless of the climate forcing pattern.

#To test this, we use a Bayesian model to simulate the INPUTS of Adams' model, river discharge (Q_in) and direct precipitation (precip). The simulated data should have the same mean and variance as the real input data, but will not follow the exact pattern that the climate did in reality. So, when we then we run these simulated draws through Adams' water balance equations, the periodicity should emerge from the physics alone.

#Note: because we are feeding predictions through a deterministic physics model, we cannot look directly at residual periodograms. We also cannot use the results of model fitting to relate input variables directly to our observed elev variable, since it is processed through another layer of equations. So, this is useful for understanding the physics of the basin, but not for the needs of our hindcast.

# Set up Adams (2015) water balance model from shared Excel spreadsheet (some help from Claude Code)
# Adams (2015) water balance data (extracted from his spreadsheet; cite Adams 2015)
adams_raw <- read.csv("Raw Data/Adams_WB_Data.csv")

# Merge Adams data with climate data (clim already loaded)
wb <- adams_raw %>%
  mutate(Calendar_Year = as.numeric(Year)) %>%
  left_join(clim, by = "Calendar_Year")
wb$Precip_cm <- wb$Precip_km * 1e5  # scale to cm/yr because models failed to converge with wildly different number sizes of predicted variables

# Area (km2) from volume (km3) — 4th order polynomial (Adams 2015)
adams_area <- function(V) {
  -0.2772258143 * V^4 + 8.5611176853 * V^3 -
    89.7086376395 * V^2 + 495.4302443649 * V + 148.4974677424
}

# Elevation (m asl) from volume (km3) — cubic (Adams 2015)
adams_elev <- function(V) {
  0.0058131042 * V^3 - 0.1668677754 * V^2 + 2.0348749694 * V + 56.3610923901
}

# Run Adams' water balance forward for a single draw
run_adams_wb <- function(Q_in_series, Precip_series = NULL, V_init = 1.98,
                         evap_range = c(145, 155)) {
  n <- length(Q_in_series)
  V <- numeric(n)
  V_prev <- V_init
  for (t in seq_len(n)) {
    A <- adams_area(V_prev)
    A <- max(A, 100)
    E <- runif(1, evap_range[1], evap_range[2]) / 100000  # cm/yr -> km/yr
    P <- if (!is.null(Precip_series)) Precip_series[t] else 0
    dV <- Q_in_series[t] + P * A - E * A
    V[t] <- max(V_prev + dV, 0.01)
    V_prev <- V[t]
  }
  adams_elev(V)
}


# Multivariate model: jointly predict Q_in and Precip from PDO + ENSO
wb_pdo <- wb %>%
  filter(!is.na(Q_in), !is.na(Precip_cm), !is.na(PDO), !is.na(ENSO))

model_path <- "Models/mv_pdo.rds"
if (file.exists(model_path)) {
  mv_pdo <- readRDS(model_path)
} else {

mv_pdo <- brm(bf(Q_in ~ PDO + ENSO) +
              bf(Precip_cm ~ PDO + ENSO) +
              set_rescor(TRUE), #this tells a multivariate model that the variables being predicted are correlated
              data = wb_pdo,
              family = gaussian(),
              prior = c(prior(normal(1.5, 1), class = "Intercept", resp = "Qin"),
                        prior(normal(0, 1), class = "b", resp = "Qin"),
                        prior(student_t(3, 0, 1), class = "sigma", resp = "Qin"),
                        prior(normal(20, 10), class = "Intercept", resp = "Precipcm"),
                        prior(normal(0, 10), class = "b", resp = "Precipcm"),
                        prior(student_t(3, 0, 5), class = "sigma", resp = "Precipcm"),
                        prior(lkj(2), class = "rescor")),
              chains = 4,
              iter = 2000,
              warmup = 1000,
              seed = 2025)

  saveRDS(mv_pdo, file = model_path)
}

summary(mv_pdo)


# Posterior predictions. Using the same draws preserves Q_in and Precip correlation
qin_pred_pdo <- posterior_predict(mv_pdo, newdata = wb_pdo, resp = "Qin")
precip_pred_pdo_cm <- posterior_predict(mv_pdo, newdata = wb_pdo, resp = "Precipcm")
qin_pred_pdo[qin_pred_pdo < 0.01] <- 0.01
precip_pred_pdo_cm[precip_pred_pdo_cm < 0] <- 0
precip_pred_pdo <- precip_pred_pdo_cm / 1e5  # back to km/yr for Adams physics

# Run 500 draws through Adams' water balance
n_draws_wb <- 500
set.seed(2025)
draw_idx_wb <- sample(nrow(qin_pred_pdo), n_draws_wb)

years_wb <- wb_pdo$Calendar_Year
obs_wb <- wb_pdo %>% select(Calendar_Year, elev) %>% filter(!is.na(elev))

elev_sim_wb <- matrix(NA, nrow = n_draws_wb, ncol = ncol(qin_pred_pdo))
for (i in seq_len(n_draws_wb)) {
  elev_sim_wb[i, ] <- run_adams_wb(qin_pred_pdo[draw_idx_wb[i], ],
                                   precip_pred_pdo[draw_idx_wb[i], ])
}

## Make a plot to visualize results.
sim_long_wb <- as.data.frame(elev_sim_wb) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "year_idx", values_to = "elev_sim") %>%
  mutate(year_idx = as.integer(gsub("V", "", year_idx)),
         Calendar_Year = years_wb[year_idx])

fig6_wb_sims <- ggplot() +
  geom_line(data = sim_long_wb,
            aes(x = Calendar_Year, y = elev_sim, group = draw,
                color = "Simulated draws"),
            alpha = 0.01, linewidth = 0.3) +
  geom_line(data = obs_wb,
            aes(x = Calendar_Year, y = elev, color = "Observed"),
            linewidth = 0.4) +
  scale_color_manual(name = NULL,
    values = c("Simulated draws" = col_wb_draw,
               "Observed"        = col_observed)) +
  guides(color = guide_legend(override.aes = list(
    alpha = 1, linewidth = c(0.5, 0.5)))) +
  labs(x = "Calendar Year AD", y = "Water Surface Elevation (masl)",
       title = "Water Balance Simulations vs Observed",
       subtitle = paste0(n_draws_wb,
                         " posterior draws: (Q_in, Precip) ~ PDO + ENSO -> Adams physics -> elevation")) +
  theme_paashi()
fig6_wb_sims
save_fig(fig6_wb_sims, "Figure_7_WB_Simulations.tiff", width_mm = 190, height_mm = 100)

#Look at periodograms (Assisted by Claude Code)
obs_wb_ts <- ts(obs_wb$elev, frequency = 1)
obs_wb_spec <- spectrum(obs_wb_ts, spans = c(5, 7), plot = FALSE)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(1/obs_wb_spec$freq, obs_wb_spec$spec, type = "l", log = "xy",
     xlab = "Period (years)", ylab = "Power",
     main = "Observed elevation", col = col_observed, lwd = 1.5)

for (j in 1:3) {
  draw_ts <- ts(elev_sim_wb[j, ], frequency = 1)
  draw_spec <- spectrum(draw_ts, spans = c(5, 7), plot = FALSE)
  plot(1/draw_spec$freq, draw_spec$spec, type = "l", log = "xy",
       xlab = "Period (years)", ylab = "Power",
       main = paste("WB sim draw", j), col = col_wb_draw, lwd = 1.5)
}

# The simulated periodograms show smooth bumps in the 32-64yr band. This confirms the periodicity is an emergent property of the basin's Area-Volume feedback, not just a feature of the climate forcing

## For our hindcast, we want to directly predict the lake's end state from SSTs, not obscure it through physics, so we use ARMA to help approximate the persistence in lake elevations, which helps inform our CI width. The periodicity inherent to the lake's physics is contained within the CIs of our hindcast.

```

## Using Models to Predict Reality
```{r predictions}
## Now that we understand what our model can tell us and what our CIs represent, let's predict. First an in-sample check against mld_SDPSarma, where ARMA has real observations to work with; then the hindcast, using the weighted pool.
#there are several ways to do a posterior prediction. posterior_predict() runs a full simulation based on a fitted model, with thousands of draws, outputting every result. This is usually used to compare a model to actual observed results. predict() runs posterior_predict, but then just outputs the mean and upper/lower 95; it is most useful if you just want to go straight to plotting results. posterior_epred ("expected prediction") runs a simulation based on a fitted model, gets the mean, then makes predictions based on the mean. This is typically used to make posterior predictions on large sets of new data, since it is significantly lighter than running a full posterior_predict, given that posterior_epred() is just relying on the mean of an already fitted model, whereas posterior_predict() or predict() would run thousands of predictions on all the new data, then return the mean.Lastly, fitted() runs posterior_epred() and then summarizes it, much as predict does for posterior_predict.

#First, let's use predict() to visualize how good the model is when arma has observations to work with. This will not have new data, it is just to compare the model against observations

as.data.frame(predict(mld_SDPSarma)) -> SDPSarma_Predict #outputs a data frame

#add time and observed data
SDPSarma_Predict$year_bp <- mld_SDPSarma$data$Year_B2005
SDPSarma_Predict$elev_obs <- mld_SDPSarma$data$elev

#create a csv that we can load later for plotting, rather than running this again
write.csv(SDPSarma_Predict, "Processed Data/SDPSarma_Predict.csv", row.names = FALSE)
#for later:
read.csv("Processed Data/SDPSarma_Predict.csv") -> SDPSarma_Predict

##Let's visualize the results quickly. We will do more in depth plotting later
#ribbon = 95% prediction interval, dark line = posterior mean, gray = observed
figS3_in_sample <- ggplot(SDPSarma_Predict, aes(x = year_bp - BP_OFFSET)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5, fill = "95% prediction interval"),
              alpha = 0.5) +
  geom_line(aes(y = elev_obs, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = Estimate, color = "Posterior mean"), linewidth = 0.5) +
  scale_color_manual(name = NULL,
    values = c("Observed" = "gray70", "Posterior mean" = col_model_line)) +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_inner)) +
  scale_x_reverse(labels = comma_format()) +
  labs(x = "Years Before Present (cal yr BP)", y = "Lake Elevation (masl)",
       title = "In-Sample Fit: ARMA with Observed Lagged Elevations",
       subtitle = "Posterior mean tracks observed; 95% PI shaded") +
  theme_paashi()
figS3_in_sample
save_fig(figS3_in_sample, "Figure_S4_InSample_ARMA.tiff", height_mm = 90)

##now let's run it back through time using the weighted pool from uncertainty propagation.

#stop the hindcast at the oldest real measurement from the shortest SST record
record_end   <- sapply(SDPS, function(v) max(clim$Year_B2005[!is.na(clim[[v]])]))
print(record_end)
HINDCAST_END <- min(record_end)
#predict for rows that have our relevant data
pred_rows <- which(!is.na(clim$Year_B2005) &
                  !is.na(clim$Seki_Int) & !is.na(clim$Davis5_Int) &
                  !is.na(clim$Pahnke_Int) & !is.na(clim$Stott_Int) &
                  clim$Year_B2005 <= HINDCAST_END)

## weighted mean, 95% CI for the mean (Fit_), and 95% prediction interval (Pred_)
set.seed(2025)
hindcast <- predict_weighted(pred_rows)
hindcast$year_bp  <- clim$Year_B2005[pred_rows]
hindcast$elev_obs <- clim$elev[pred_rows]

write.csv(hindcast, "Processed Data/Hindcast_Full.csv", row.names = FALSE)
hindcast <- read.csv("Processed Data/Hindcast_Full.csv")

# Rebuild Model_Outputs_Combined.csv from updated hindcast + archaeological data
arch_raw <- read.csv("Processed Data/Lake12All.csv")
combo_rebuilt <- hindcast %>%
  rename(year_b2005 = year_bp,
         Fitted_Estimate = Fit_Est, Fitted_Est.Error = Fit_SE,
         Fitted_Q2.5 = Fit_Q2.5, Fitted_Q97.5 = Fit_Q97.5,
         Predict_Estimate = Pred_Est, Predict_Est.Error = Pred_SE,
         Predict_Q2.5 = Pred_Q2.5, Predict_Q97.5 = Pred_Q97.5) %>%
  mutate(calendar_year = 2005 - year_b2005) %>%
  select(year_b2005, calendar_year, Fitted_Estimate, Fitted_Est.Error,
         Fitted_Q2.5, Fitted_Q97.5, Predict_Estimate, Predict_Est.Error,
         Predict_Q2.5, Predict_Q97.5, elev_obs) %>%
  left_join(arch_raw %>% select(ybp, elev) %>% rename(year_b2005 = ybp, arch_model = elev),
            by = "year_b2005")
write.csv(combo_rebuilt, "Processed Data/Model_Outputs_Combined.csv", row.names = FALSE)

# two-ribbon plot: outer = prediction interval, inner = credible interval for mean
ggplot(hindcast, aes(x = year_bp - BP_OFFSET)) +
  geom_ribbon(aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5),
              fill = col_model_ribbon, alpha = 0.3) +
  geom_ribbon(aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5),
              fill = col_model_inner, alpha = 0.5) +
  geom_line(aes(y = Fit_Est), color = col_model_line, linewidth = 0.5) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.5,
            alpha = 0.7, na.rm = TRUE) +
  labs(x = "Years Before Present (cal yr BP)", y = "Lake Elevation (masl)",
       title = "Lake Elevation Hindcast: Expected Trend and Prediction Intervals",
       subtitle = "Inner band (dark) = 95% CI for mean; Outer band (light) = 95% prediction interval; Dark = observed") +
  theme_paashi()

```

## Publication Figures

### Full Hindcast Figure (assisted by Claude Code)
```{r fig4_hindcast, fig.width = 7.5, fig.height = 5}
hindcast <- read.csv("Processed Data/Hindcast_Full.csv")
#convert to cal yr BP for plotting; the csv stores years before 2005
hindcast$year_bp <- hindcast$year_bp - BP_OFFSET
training_boundary <- 2006 - BP_OFFSET
pleis_holo        <- 11700

# Panel A: Holocene (0-12,000 cal yr BP)
hindcast_holo <- hindcast %>% filter(year_bp <= 12000)
hindcast_holo_model <- hindcast_holo %>% filter(year_bp > training_boundary)
hindcast_model <- hindcast %>% filter(year_bp > training_boundary)

fig3a <- ggplot(hindcast_holo, aes(x = year_bp)) +
  annotate("rect", xmin = -BP_OFFSET, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(data = hindcast_holo_model,
              aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5, fill = "95% prediction interval"), alpha = 0.4) +
  geom_ribbon(data = hindcast_holo_model,
              aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5, fill = "95% CI (expected trend)"), alpha = 0.5) +
  geom_line(data = hindcast_holo_model,
            aes(y = Fit_Est, color = "Model trend"), linewidth = 0.4) +
  geom_line(aes(y = elev_obs, color = "Observed"), linewidth = 0.3,
            alpha = 0.7, na.rm = TRUE) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.5) +
  annotate("text", x = training_boundary + 200, y = 69,
           label = "Calibration
boundary", size = 3.0, hjust = 1, color = "gray40") +
  annotate("text", x = pleis_holo - 200, y = 69,
           label = "Pleistocene-
Holocene
boundary", size = 3.0,
           hjust = 0, color = col_pleis_line) +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_ribbon,
               "95% CI (expected trend)" = col_model_inner)) +
  scale_color_manual(name = NULL,
    values = c("Model trend" = col_model_line,
               "Observed" = col_observed)) +
  scale_x_reverse(labels = comma_format(),
                  breaks = seq(0, 12000, by = 2000)) +
  coord_cartesian(ylim = c(50, 72)) +
  labs(x = NULL, y = "Lake Elevation (masl)",
       title = "A. Holocene (0–12,000 cal yr BP)") +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme_paashi() +
  theme(legend.position = "none")

fig3b <- ggplot(hindcast, aes(x = year_bp)) +
  annotate("rect", xmin = -BP_OFFSET, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(data = hindcast_model,
              aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5, fill = "95% prediction interval"), alpha = 0.4) +
  geom_ribbon(data = hindcast_model,
              aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5, fill = "95% CI (expected trend)"), alpha = 0.5) +
  geom_line(data = hindcast_model,
            aes(y = Fit_Est, color = "Model trend"), linewidth = 0.4) +
  geom_line(aes(y = elev_obs, color = "Observed"), linewidth = 0.3,
            alpha = 0.7, na.rm = TRUE) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.5) +
  annotate("text", x = 18000, y = 70,
           label = "Model calibrated on Holocene data;
deeper predictions carry
greater uncertainty",
           size = 3.0, color = "gray40", fontface = "italic") +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_ribbon,
               "95% CI (expected trend)" = col_model_inner)) +
  scale_color_manual(name = NULL,
    values = c("Model trend" = col_model_line,
               "Observed" = col_observed)) +
  scale_x_reverse(labels = comma_format(),
                  breaks = seq(0, 22000, by = 4000)) +
  coord_cartesian(ylim = c(50, 72)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Lake Elevation (masl)",
       title = "B. Full reconstruction (0–20,000 cal yr BP)") +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme_paashi() +
  theme(legend.position = "bottom")

fig3 <- fig3a / fig3b +
  plot_annotation(
    title = "Bayesian Hindcast of Pa'ashi Elevation")
fig3
save_fig(fig3, "Figure_4_Hindcast.tiff", width_mm = 190, height_mm = 160)
```

### Negrini (2006) Archaeological Comparison (assisted by Claude Code)

Two-panel figure: (A) recreation based on data from Negrini et al. (2006) Fig. 10
from digitized observations, deep-water clay intervals, and trench horizons;
(B) Negrini's dashed reconstruction overlaid on our Bayesian hindcast.

```{r fig5_negrini_comparison, fig.width = 7.5, fig.height = 7.5}
# --- Data ----------------------------------------------------------------
hindcast  <- read.csv("Processed Data/Hindcast_Full.csv")
combo     <- read.csv("Processed Data/Model_Outputs_Combined.csv")
neg_horiz <- read.csv("Raw Data/Negrini_Horizons.csv",
                      stringsAsFactors = FALSE)
neg_clay  <- read.csv("Raw Data/Negrini_DeepClay.csv",
                      stringsAsFactors = FALSE) %>%
  left_join(
    neg_horiz %>% rename(h_ybp_start  = ybp_start,
                         h_ybp_end    = ybp_end,
                         h_elev_start = elev_start,
                         h_elev_end   = elev_end),
    by = "trench"
  ) %>%
  mutate(slope      = (h_elev_end - h_elev_start) /
                      (h_ybp_end - h_ybp_start),
         elev_start = h_elev_start + slope * (ybp_start - h_ybp_start),
         elev_end   = h_elev_start + slope * (ybp_end   - h_ybp_start)) %>%
  select(trench, ybp_start, ybp_end, elev_start, elev_end)

# Per-type metadata. Shapes/colors/labels are derived from this single tibble
# rather than maintained as parallel vectors.
obs_meta <- tribble(
  ~type,           ~shape, ~depth,          ~label,
  "pollen",        0,      "shallow water", "Littoral pollen",
  "organic",       3,      "shallow water", "Organic-rich (marsh?) deposits",
  "mudcracks",     6,      "shallow water", "Mudcracks",
  "beach_sands",   4,      "shallow water", "Well-sorted beach? sands",
  "pelagic_algae", 17,     "deeper water",  "Pelagic algae",
  "laminations",   15,     "deeper water",  "Laminations in silts/clays"
)
depth_colors     <- c("deeper water" = "#1F4E79", "shallow water" = "#2E7D52")
col_dc           <- "gray30"
col_dc_label     <- "Deep-water clay interval"
col_ref          <- "gray45"
rambla_highstand <- 70.1

shallow_meta <- obs_meta %>% filter(depth == "shallow water")
deeper_meta  <- obs_meta %>% filter(depth == "deeper water")

neg_obs <- read.csv("Raw Data/Negrini_Observations.csv",
                    stringsAsFactors = FALSE) %>%
  rename(year_b2005 = ybp) %>%
  mutate(type = recode(obs,
                       "silt clay"     = "laminations",
                       "organic marsh" = "organic",
                       "algae"         = "pelagic_algae",
                       "beach sands"   = "beach_sands")) %>%
  inner_join(obs_meta %>% select(type, depth), by = "type")

#convert every x value to cal yr BP for plotting (the csv files store years before 2005)
hindcast  <- hindcast  %>% mutate(year_bp = year_bp - BP_OFFSET)
combo     <- combo     %>% mutate(year_b2005 = year_b2005 - BP_OFFSET)
neg_horiz <- neg_horiz %>% mutate(ybp_start = ybp_start - BP_OFFSET, ybp_end = ybp_end - BP_OFFSET)
neg_clay  <- neg_clay  %>% mutate(ybp_start = ybp_start - BP_OFFSET, ybp_end = ybp_end - BP_OFFSET)
neg_obs   <- neg_obs   %>% mutate(year_b2005 = year_b2005 - BP_OFFSET)

shallow_pts <- neg_obs %>% filter(depth == "shallow water")
deeper_pts  <- neg_obs %>% filter(depth == "deeper water")

# Upward "lake rose at least this high" arrows. Clay-bar arrows are gray
# (same shade as the bars); deeper-water observation arrows are blue.
clay_arrows <- bind_rows(
  neg_clay %>% transmute(x = ybp_start, y = elev_start),
  neg_clay %>% transmute(x = ybp_end,   y = elev_end)
)
deep_arrows <- deeper_pts %>% transmute(x = year_b2005, y = elev)
arrow_style <- arrow(length = unit(0.08, "cm"), type = "open")

# Holocene subsets for panel B
hindcast_holo       <- hindcast %>% filter(year_bp <= 12000)
hindcast_holo_model <- hindcast_holo %>% filter(year_bp > training_boundary)
combo_holo          <- combo %>% filter(year_b2005 <= 12000)
combo_holo_model    <- combo_holo %>% filter(year_b2005 > training_boundary)

# Negrini's reconstruction line (for panel A; panel B uses combo directly)
arch_line <- combo %>%
  filter(!is.na(arch_model), year_b2005 >= -BP_OFFSET, year_b2005 <= 11500) %>%
  select(year_b2005, elev = arch_model)

# --- Panel A: Negrini Fig. 10 recreation ---------------------------------
# Layer order matters for ggnewscale: shallow-water points use the first
# shape scale; everything after `new_scale("shape")` uses the second.
fig4a <- ggplot() +
  geom_hline(yintercept = rambla_highstand, linetype = "dotted",
             color = col_ref, linewidth = 0.4) +
  annotate("text", x = 9000, y = rambla_highstand + 1.1,
           label = "Top of Rambla highstand",
           size = 3.0, color = "gray30") +
  geom_segment(data = neg_horiz,
               aes(x = ybp_start, xend = ybp_end,
                   y = elev_start, yend = elev_end),
               linetype = "dashed", color = col_ref, linewidth = 0.45) +
  annotate("text", x = 5000, y = 62.3, label = "shoreline trench",
           size = 3.0, color = "gray30", fontface = "italic") +
  annotate("text", x = 1000, y = 56.9, label = "lake plain trench",
           size = 3.0, color = "gray30", fontface = "italic") +
  annotate("text", x = 7000, y = 51.8,
           label = "depocenter core (Davis 1999)",
           size = 3.0, color = "gray30", fontface = "italic") +
  # Background bars/arrows/line — drawn BEFORE points so points sit on top.
  geom_segment(data = neg_clay,
               aes(x = ybp_start, xend = ybp_end,
                   y = elev_start, yend = elev_end),
               color = col_dc, linewidth = 1.3) +
  geom_segment(data = clay_arrows,
               aes(x = x, xend = x, y = y, yend = y + 1.6),
               arrow = arrow_style, color = col_dc, linewidth = 0.3) +
  geom_segment(data = deep_arrows,
               aes(x = x, xend = x, y = y, yend = y + 1.6),
               arrow = arrow_style,
               color = depth_colors[["deeper water"]], linewidth = 0.4) +
  geom_line(data = arch_line,
            aes(x = year_b2005, y = elev),
            color = col_arch, linewidth = 0.7, linetype = "dashed") +
  # --- Shallow-water shape scale ---
  geom_point(data = shallow_pts,
             aes(x = year_b2005, y = elev, shape = type),
             color = depth_colors[["shallow water"]],
             size = 2.4, stroke = 1.3) +
  scale_shape_manual(name = "shallow water",
                     values = setNames(shallow_meta$shape, shallow_meta$type),
                     labels = setNames(shallow_meta$label, shallow_meta$type),
                     breaks = shallow_meta$type,
                     guide = guide_legend(order = 1, nrow = 1, byrow = TRUE)) +
  new_scale("shape") +
  # --- Deeper-water + deep-clay shape scale ---
  geom_point(data = deeper_pts,
             aes(x = year_b2005, y = elev, shape = type),
             color = depth_colors[["deeper water"]],
             size = 2.4, stroke = 0.9) +
  # Phantom off-canvas point whose only job is to register the
  # "Deep-water clay interval" entry in the shape scale. (geom_segment
  # silently drops shape mappings, so we need a point geom.) key_glyph
  # forces the legend to draw a horizontal line instead of a point.
  geom_point(data = data.frame(year_b2005 = -100, elev = -100,
                               type = col_dc_label),
             aes(x = year_b2005, y = elev, shape = type),
             color = col_dc, key_glyph = "path") +
  scale_shape_manual(
    name = "deeper water",
    values = c(setNames(deeper_meta$shape, deeper_meta$type),
               setNames(15, col_dc_label)),
    labels = c(setNames(deeper_meta$label, deeper_meta$type),
               setNames(col_dc_label, col_dc_label)),
    breaks = c(deeper_meta$type, col_dc_label),
    guide = guide_legend(
      order = 2, nrow = 1, byrow = TRUE,
      override.aes = list(
        shape     = c(deeper_meta$shape, NA),
        color     = c(rep(depth_colors[["deeper water"]], nrow(deeper_meta)),
                      col_dc),
        linetype  = c(rep(0, nrow(deeper_meta)), 1),
        linewidth = c(rep(0, nrow(deeper_meta)), 1.3)
      ))) +
  scale_x_reverse(labels = comma_format(),
                  breaks = seq(0, 12000, by = 2000),
                  expand = expansion(mult = c(0.02, 0.01))) +
  coord_cartesian(xlim = c(11800, -BP_OFFSET), ylim = c(50, 72)) +
  labs(x = NULL, y = "Lake Elevation (masl)",
       title = "A. Negrini et al. (2006) Fig. 10 recreation",
       subtitle = paste0("Holocene lake-level history from digitized ",
                         "observations and trench horizons.")) +
  theme_paashi() +
  theme(legend.position = "bottom",
        legend.box = "vertical")

# --- Panel B: simplified Bayesian vs Negrini comparison ------------------
fig4b <- ggplot() +
  annotate("rect", xmin = -BP_OFFSET, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(data = hindcast_holo_model,
              aes(x = year_bp, ymin = Pred_Q2.5, ymax = Pred_Q97.5,
                  fill = "95% prediction interval"), alpha = 0.5) +
  geom_line(data = combo_holo_model,
            aes(x = year_b2005, y = Fitted_Estimate,
                color = "Bayesian model trend"),
            linewidth = 0.4, alpha = 0.85) +
  geom_line(data = combo_holo %>% filter(!is.na(elev_obs)),
            aes(x = year_b2005, y = elev_obs,
                color = "Observed"),
            linewidth = 0.25, alpha = 0.55) +
  geom_line(data = combo_holo %>% filter(!is.na(arch_model)),
            aes(x = year_b2005, y = arch_model,
                color = "Negrini reconstruction"),
            linewidth = 0.7, linetype = "dashed", alpha = 0.9) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_ribbon),
    guide = guide_legend(order = 1)) +
  scale_color_manual(name = NULL,
    values = c("Bayesian model trend"   = col_model_line,
               "Negrini reconstruction" = col_arch,
               "Observed"               = col_observed),
    guide = guide_legend(order = 2,
                         override.aes = list(
                           linetype  = c("solid", "dashed", "solid"),
                           linewidth = c(0.5,     0.7,      0.4)))) +
  scale_x_reverse(labels = comma_format(),
                  breaks = seq(0, 12000, by = 2000),
                  expand = expansion(mult = c(0.02, 0.01))) +
  coord_cartesian(xlim = c(11800, -BP_OFFSET), ylim = c(48, 73)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Lake Elevation (masl)",
       title = "B. Bayesian hindcast vs Negrini reconstruction") +
  theme_paashi() +
  theme(legend.position = "bottom")

fig4 <- fig4a / fig4b +
  plot_annotation(title = "Comparison with Negrini et al. (2006) Reconstruction")
fig4
save_fig(fig4, "Figure_5_Negrini_Comparison.tiff",
         width_mm = 190, height_mm = 200)
```

## Session Info
```{r session_info}
sessionInfo()
```
