---
title: "Paashi_Paleo_Models_Pub"
author: "Robin_ML"
date: "2025-12-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages and Raw Data
```{r loading packages/data}

library(tidyverse)
library(tidybayes)
library(ggplot2)
library(ggthemes)
library(brms)
library(bayesplot)
library(loo)
library(patchwork)
library(scales)
library(ggrepel)
library(maps)

clim <- read.csv("Raw Data/ElevSSTs.csv")

```

## Publication Theme and Palette (assisted by Claude Code)
```{r theme_and_palette}
# Colorblind-safe, grayscale-distinguishable palette
col_model_line   <- "#1F4E79"   # dark blue — model trend line
col_model_ribbon <- "#A8C8E8"   # light blue — prediction interval
col_model_inner  <- "#4682B4"   # steel blue — fitted CI
col_observed     <- "gray20"    # near-black — observed data
col_arch         <- "#B85C38"   # terra cotta — archaeological reconstruction
col_seki         <- "#D4930D"   # amber — Seki SST (California Margin)
col_palmer       <- "#1B998B"   # teal — Palmer SST (W. Equatorial Pacific)
col_davis5       <- "#5B7B9A"   # slate blue — Davis5 SST (Sea of Okhotsk)
col_train_bg     <- "#FFF8F0"   # faint warm — training window background
col_pleis_line   <- "gray50"    # Pleistocene-Holocene boundary

# Shared theme for JAS (Elsevier) publication — min 7pt text, sans-serif
theme_paashi <- function(base_size = 8) {
  theme_minimal(base_size = base_size, base_family = "sans") %+replace%
    theme(
      plot.title       = element_text(size = rel(1.15), face = "bold",
                                      hjust = 0, margin = margin(b = 4)),
      plot.subtitle    = element_text(size = rel(0.95), color = "gray30",
                                      hjust = 0, margin = margin(b = 6)),
      axis.title       = element_text(size = rel(1.0)),
      axis.text        = element_text(size = rel(0.9), color = "gray20"),
      strip.text       = element_text(size = rel(1.0), face = "bold"),
      legend.text      = element_text(size = rel(0.9)),
      legend.title     = element_text(size = rel(0.95), face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
      plot.margin      = margin(8, 8, 8, 8),
      legend.position  = "bottom",
      legend.key.size  = unit(0.4, "cm")
    )
}

# Save figures at JAS dimensions
save_fig <- function(plot, filename,
                     width_mm = 190, height_mm = 120,
                     dpi = 500, device = "tiff") {
  ggsave(
    filename = file.path("Figures", filename),
    plot     = plot,
    width    = width_mm, height = height_mm,
    units    = "mm", dpi = dpi, device = device,
    compression = if (device == "tiff") "lzw" else NULL
  )
}

dir.create("Figures", showWarnings = FALSE)
```

## Checking Paleoclimate Data
```{r plotting}

par(mfrow = c(3, 2))
plot(clim$Year_B2005, clim$Seki_Int, main = "Seki (NE Pacific, 28kya, ~400yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Palmer_Int, main = "Palmer (W Eq Pacific, 23kya, ~600yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Davis1, main = "Davis-1/Barron (NE Pacific, 15kya, ~100yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Davis5, main = "Davis-5 (Okhotsk, 25kya, ~300yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Isono_Int, main = "Isono (Japan, 15kya, ~20yr res) - EXCLUDED", pch = ".", xlim = c(0, 25000))
par(mfrow = c(1, 1))


##Examine training data
ggplot(clim, aes(x = elev)) +
  geom_histogram(bins = 300)
#This gives a normal distribution, so no further action needed; we can use family = gaussian. The distribution may in fact be bimodal with two very close peaks near the center-- this may represent "drought" and "flood" periods, as truly average rainfall is rarity in California according to oral traditions. However, these peaks are close enough to treat as a single normal distribution; bimodal modeling produces bad results

```

## Define Explicit Priors
```{r define_priors}
## Inspect what brms would choose as defaults for our final model
default_priors <- get_prior(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
  data = clim,
  family = gaussian())

print(default_priors)

## Define weakly informative priors in case others have different versions of R or BRMS
# Intercept: Normal(63, 5) -- centered on mean observed lake elevation (~63 MaSL), 95% prior mass between 53-73m, generous given basin physical limits. Modern basin bottom is at 55masl, and highest visible geological shoreline is ~70masl. Note that over time, lake basins may rise due to sediment.
# Slopes: Normal(0, 5) -- 1-degree SST change shifts lake by <10m (2 SD);  generous since entire observed elev range is ~9m
# Sigma: half-Student-t(3, 0, 3) -- standard weakly informative for residual SD
# AR/MA: brms defaults to uniform(-1, 1)

# Base priors for models without autocorrelation terms
paashi_priors <- c(
  prior(normal(63, 5), class = "Intercept"),
  prior(normal(0, 5), class = "b"),
  prior(student_t(3, 0, 3), class = "sigma"))

# Extended priors for ARMA models (ar/ma must be explicit for prior predictive checks)
paashi_priors_arma <- c(
  paashi_priors,
  prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1),
  prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1))

```

## Prior Predictive Check
```{r prior_predictive_check}
# Fit model with sample_prior = "only" to draw from priors without seeing data to verify our priors are weakly informative but plausible
mld_prior_only <- brm(
  elev ~ Seki_Int + Palmer_Int + Davis5 +
    arma(p = 1, q = 1, time = Year_B2005),
  data = clim,
  family = gaussian(),
  prior = paashi_priors_arma,
  sample_prior = "only",
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2025)

# Simulated data should include the observed range but not be too close, as these priors are weakly informative
pp_check(mld_prior_only, ndraws = 100) +
  labs(title = "Prior Predictive Check",
       subtitle = "Dark = observed; light = simulated from prior only") +
  xlim(30, 100)
## looks good!

```

## Create and Test Models
```{r model testing}
### running brms on paleoclimate datasets
##model sea surface temperature relationships to elevation


##Single-Predictor Models
# Test each SST record individually

# Seki (NE Pacific / California Margin, longest record)
mld_Seki <- brm(elev ~ Seki_Int,
                data = clim,
                family = gaussian(),
                prior = paashi_priors,
                chains = 4, iter = 3000, warmup = 1000,
                seed = 2025)
summary(mld_Seki)

# Palmer (Western Equatorial Pacific, ENSO-related signal)
mld_Palmer <- brm(elev ~ Palmer_Int,
                  data = clim,
                  family = gaussian(),
                  prior = paashi_priors,
                  chains = 4, iter = 3000, warmup = 1000,
                  seed = 2025)
summary(mld_Palmer)

# Davis1/Barron (NE Pacific, highest resolution among usable records)
mld_Davis1 <- brm(elev ~ Davis1,
                  data = clim,
                  family = gaussian(),
                  prior = paashi_priors,
                  chains = 4, iter = 3000, warmup = 1000,
                  seed = 2025)
summary(mld_Davis1)

# Davis5 (Sea of Okhotsk, long record with good resolution)
mld_Davis5 <- brm(elev ~ Davis5,
                  data = clim,
                  family = gaussian(),
                  prior = paashi_priors,
                  chains = 4, iter = 3000, warmup = 1000,
                  seed = 2025)
summary(mld_Davis5)

## Compare Single-Predictor Models
loo(mld_Seki, mld_Palmer, mld_Davis1, mld_Davis5, compare = TRUE)
#Model comparisons:
#           elpd_diff se_diff
#mld_Davis5   0.0       0.0  
#mld_Davis1  -1.9       7.5  
#mld_Palmer  -5.8       5.8  
#mld_Seki   -21.0       6.2


## Combined-Predictor Models

# mld_Long: Three longest records (for extended hindcast to ~23kya)
# Seki (28kya) + Palmer (23kya) + Davis5 (25kya)
mld_Long <- brm(elev ~ Seki_Int + Palmer_Int + Davis5,
                data = clim,
                family = gaussian(),
                prior = paashi_priors,
                chains = 4, iter = 3000, warmup = 1000,
                seed = 2025)
summary(mld_Long)
saveRDS(mld_Long, file = "Models/mld_Long.rds")

# mld_Short: All four records (best fit, limited to ~15kya by Davis1)
# Adds Davis1/Barron for improved fit within Holocene
mld_Short <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + Davis1,
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors,
                 chains = 4, iter = 3000, warmup = 1000,
                 seed = 2025)
summary(mld_Short)


#model 2 closest to California: Seki (NE Pacific "California Margin"), Davis1/Barron (California Margin)
mld_Cal <- brm(elev ~ Seki_Int + Davis1,
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors,
                 chains = 4,
                 iter = 3000,
                 warmup = 1000,
                 seed = 2025)
summary(mld_Cal)


## Compare Combined Models
loo(mld_Long, mld_Short, mld_Cal, compare = TRUE)
#Model comparisons:
#          elpd_diff se_diff
#mld_Short   0.0       0.0  
#mld_Long  -15.8       5.6  
#mld_Cal   -26.1       7.3

## Based off of these results, mld_Long looks like the way to go. It is not too terrible compared to mld_Short, and will allow us to predict much further back in time.

```

### Figure 1: Study Area and Data Overview (assisted by Claude Code)
```{r fig1_map_and_sst, fig.width = 7.5, fig.height = 6}

##Let's visualize where our selected SSTs are coming from and what they look like.

# Panel A: Pacific Basin map with SST core sites + Tulare Lake
sites <- data.frame(
  name = c("Tulare Lake", "Seki et al.\n(California Margin)",
           "Palmer & Pearson\n(W. Equatorial Pacific)",
           "Davis et al.\n(Sea of Okhotsk)"),
  lon  = c(-119.5, -126.0, 159.0, 145.0),
  lat  = c(35.9, 42.0, 0.0, 50.0),
  type = c("lake", "sst", "sst", "sst"),
  col  = c(col_observed, col_seki, col_palmer, col_davis5)
)

world2 <- map_data("world2")
sites$lon360 <- ifelse(sites$lon < 0, sites$lon + 360, sites$lon)

fig1a <- ggplot() +
  geom_polygon(data = world2, aes(x = long, y = lat, group = group),
               fill = "gray92", color = "gray70", linewidth = 0.15) +
  geom_point(data = sites, aes(x = lon360, y = lat),
             color = sites$col, fill = sites$col,
             shape = c(18, 16, 16, 16), size = c(3, 2.5, 2.5, 2.5)) +
  geom_text_repel(data = sites, aes(x = lon360, y = lat, label = name),
                  size = 2.2, color = sites$col,
                  box.padding = 0.5, max.overlaps = 10,
                  fontface = "bold", lineheight = 0.85) +
  coord_cartesian(xlim = c(100, 300), ylim = c(-15, 58), expand = FALSE) +
  scale_x_continuous(breaks = c(120, 150, 180, 210, 240, 270),
                     labels = c("120\u00B0E", "150\u00B0E", "180\u00B0",
                                "150\u00B0W", "120\u00B0W", "90\u00B0W")) +
  labs(title = "A. Study area and paleoclimate data sources",
       x = NULL, y = NULL) +
  theme_paashi() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 6),
        panel.grid.major = element_line(color = "gray85", linewidth = 0.2))

# Panel B: SST predictor time series
sst_interp <- clim %>%
  select(Year_B2005, Seki_Int, Palmer_Int, Davis5) %>%
  pivot_longer(-Year_B2005, names_to = "source", values_to = "sst_interp") %>%
  mutate(source = recode(source,
    "Seki_Int"   = "Seki (California Margin)",
    "Palmer_Int" = "Palmer (W. Equatorial Pacific)",
    "Davis5"     = "Davis (Sea of Okhotsk)"))

sst_raw <- clim %>%
  select(Year_B2005, Seki_SST, Palmer_SST) %>%
  pivot_longer(-Year_B2005, names_to = "source", values_to = "sst_raw") %>%
  filter(!is.na(sst_raw)) %>%
  mutate(source = recode(source,
    "Seki_SST"  = "Seki (California Margin)",
    "Palmer_SST" = "Palmer (W. Equatorial Pacific)"))

sst_colors <- c(
  "Seki (California Margin)"       = col_seki,
  "Palmer (W. Equatorial Pacific)"  = col_palmer,
  "Davis (Sea of Okhotsk)"          = col_davis5)

fig1b <- ggplot() +
  geom_line(data = sst_interp,
            aes(x = Year_B2005, y = sst_interp, color = source),
            linewidth = 0.3, alpha = 0.7) +
  geom_point(data = sst_raw,
             aes(x = Year_B2005, y = sst_raw, color = source),
             size = 0.4, alpha = 0.5) +
  scale_color_manual(values = sst_colors) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 28000, by = 4000)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Sea Surface Temperature (\u00B0C)",
       title = "B. Pacific SST records used as predictors",
       subtitle = "Lines = interpolated; points = original measurements",
       color = NULL) +
  theme_paashi() +
  theme(legend.position = "bottom")

fig1 <- fig1a / fig1b + plot_layout(heights = c(3, 3))
fig1
## save_fig(fig1, "Figure_1_Data_Overview.tiff", width_mm = 190, height_mm = 170)
```

##Add Prior Year Influence
```{r refining_models}
##Now that we have selected our SST sets with the best correlation to lake elevation, we want to see if other factors play in. Because this is a hydrological system, it is likely that the prior year can influence the next year.

##first, let's check our "auto correlation function" or ACF, to see how correlated each year's value is to the prior year; this will help us design better models.

#fix data for acf and pacf
clim <- clim[order(clim$Year_B2005), ] #make double sure clim is read in order of year
y <- clim$elev
y_complete <- stats::na.omit(y) #acf and pacf will not work if there are NAs

# acf checks auto correlation to see how long a value can influence subsequent values. PACF (partial ACF) removes the overlap so you can see more clearly the direct influence of prior years.
acf(y_complete, main = "ACF of elev (NAs dropped)")
pacf(y_complete, main = "PACF of elev (NAs dropped)")

#looks like we are strongly correlated for 1 year prior in the PACF. So, in ACF, value is strongly correlated with prior values, slowly tapering until ~15 prior values.
#periodicity around ~40 years backs up oral traditions, but is not necessarily strong enough to warrant adding fourier terms or seasonal arma. We will investigate periodicity in a bit.

## now we can run our best elev ~ SST model, adding influence from prior years. There are two arguments for adding in auto-correlation to a model. ar(p=x) says that the current value is correlated with past VALUES of x prior years. ma(q=x) says that the current value is correlated with past SHOCKS, or large CHANGES in value from the expected/mean value, from x prior years. arma(p = , q = ) just combines both. We can determine our p and q values from our ACF and PACF results.

#because we have a big spike in correlation for the previous year, plus a small correlation with the prior year in PACF; and a smooth taper in ACF, we can see that what matters most strongly is the VALUE of the prior year. If there were multiple spikes before taper in ACF, then we might pay more attention to our q value for ma().

#given what we see here, we can probably be safe with ar(p = 1), but could also reasonably try arma(p = 1, q = 1). 

##model Long with previous year effects
#ar(p = 1) for long, smooth persistence, good in hydrological models
mld_Longar <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + ar(p = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = c(paashi_priors,
                           prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1)),
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longar)
#plot(mld_Longar)
#pp_check(mld_Longar, ndraws = 100)


#ma(q = 1) does the deviation from the mean matter more than the actual value? Let's check ma
mld_Longma <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + ma(q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = c(paashi_priors,
                           prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1)),
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longma)
#plot(mld_Longma)
#pp_check(mld_Longma, ndraws = 100)

#arma(p = 1, q = 1) combines to see if both matter.
mld_Longarma <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longarma)
plot(mld_Longarma)
pp_check(mld_Longarma, ndraws = 100)
saveRDS(mld_Longarma, file = "Models/mld_Longarma.rds")
readRDS("Models/mld_Longarma.rds") -> mld_Longarma

loo(mld_Long, mld_Longar, mld_Longma, mld_Longarma, compare = TRUE)
# Results:
#              elpd_diff se_diff
# mld_Longarma    0.0       0.0 
# mld_Longar     -4.1       3.4 
# mld_Longma   -405.7      30.9 
# mld_Long     -998.7      42.5

##Longarma is best!

```

### Model Diagnostics Figure (assisted by Claude Code)
```{r figS1_diagnostics, fig.width = 7.5, fig.height = 8}
mld_Longarma <- readRDS("Models/mld_Longarma.rds")

# Panel A: Trace plots
s1a <- mcmc_trace(mld_Longarma,
                  pars = c("b_Intercept", "b_Seki_Int", "b_Palmer_Int",
                           "b_Davis5", "sigma", "ar[1]", "ma[1]"),
                  facet_args = list(ncol = 1)) +
  labs(title = "A. Trace plots") +
  theme_paashi(base_size = 7)

# Panel B: Posterior predictive density overlay
s1b <- pp_check(mld_Longarma, ndraws = 100, type = "dens_overlay") +
  labs(title = "B. Posterior predictive check") +
  theme_paashi(base_size = 7)

# Panel C: ACF posterior predictive check
yrep   <- posterior_predict(mld_Longarma, ndraws = 200)
y_obs  <- mld_Longarma$data$elev
acf_obs_vals <- acf(y_obs[!is.na(y_obs)], lag.max = 20, plot = FALSE)$acf[,,1]

acf_reps <- apply(yrep[1:200, !is.na(y_obs)], 1, function(y) {
  acf(y, lag.max = 20, plot = FALSE)$acf[,,1]
})

acf_rep_df <- as.data.frame(t(acf_reps)) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "lag", values_to = "acf") %>%
  mutate(lag = as.integer(gsub("V", "", lag)) - 1)

acf_obs_df <- data.frame(lag = 0:20, acf = acf_obs_vals)

s1c <- ggplot() +
  geom_line(data = acf_rep_df, aes(x = lag, y = acf, group = draw),
            alpha = 0.08, color = "gray50") +
  geom_line(data = acf_obs_df, aes(x = lag, y = acf),
            color = "black", linewidth = 0.8) +
  labs(x = "Lag (years)", y = "Autocorrelation",
       title = "C. Posterior predictive ACF check",
       subtitle = "Black = observed; gray = simulated") +
  theme_paashi(base_size = 7)

# Panel D: LOO Pareto k diagnostics
loo_longarma <- loo(mld_Longarma)
k_values <- loo_longarma$diagnostics$pareto_k
k_df <- data.frame(obs = seq_along(k_values), k = k_values)

s1d <- ggplot(k_df, aes(x = obs, y = k)) +
  geom_point(size = 0.3, alpha = 0.4, color = col_model_inner) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange",
             linewidth = 0.4) +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "red",
             linewidth = 0.4) +
  annotate("text", x = nrow(k_df) * 0.85, y = 0.55,
           label = "k = 0.5 (caution)", size = 2.2, color = "orange") +
  annotate("text", x = nrow(k_df) * 0.85, y = 0.75,
           label = "k = 0.7 (unreliable)", size = 2.2, color = "red") +
  labs(x = "Observation", y = "Pareto k",
       title = "D. LOO-CV Pareto k diagnostics",
       subtitle = paste0("All ", nrow(k_df), " observations below 0.5")) +
  theme_paashi(base_size = 7)

figS1 <- (s1a | (s1b / s1c / s1d)) +
  plot_layout(widths = c(1, 1)) +
  plot_annotation(
    title = "Supplementary Figure S1: Model Diagnostics (mld_Longarma)")
figS1
## save_fig(figS1, "Figure_S1_Diagnostics.tiff", width_mm = 190, height_mm = 220)
```

## Continue Model Testing
```{r model_testing_continued}
# Test whether SSTs are necessary, or if ARMA alone suffices
# No "b" class here since there are no slope predictors (intercept-only)
mld_arma_only <- brm(elev ~ 1 + arma(p = 1, q = 1, time = Year_B2005),
                     data = clim,
                     family = gaussian(),
                     prior = c(
                       prior(normal(63, 5), class = "Intercept"),
                       prior(student_t(3, 0, 3), class = "sigma"),
                       prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1),
                       prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1)),
                     chains = 4, iter = 4000, warmup = 2000,
                     seed = 2025,
                     control = list(adapt_delta = 0.95))

loo(mld_Longarma, mld_arma_only, compare = TRUE)
# Results: elpd_diff = -0.1 (SE 2.5) -- essentially identical for year-to-year prediction. This is expected in loo since ARMA dominates year over year correlation; loo (leave one out) observes prediction when one observation is left out at a time. SSTs are essential for century+ scale climate forcing, because arma alone does not inform the mean when making predictions on new data.

##let's check for non-linear and more complex relationships

#spline instead of linear
mld_Longarmas <- brm(elev ~ s(Seki_Int) + s(Palmer_Int) + s(Davis5) + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longarmas)
#plot(mld_Longarmas)
#pp_check(mld_Longarmas, ndraws = 100)
saveRDS(mld_Longarmas, file = "Models/mld_Longarmas.rds")

#linear interactions between variables
# Using * for full interaction hierarchy (main effects + 2-way + 3-way)
mld_Longarmar <- brm(elev ~ Seki_Int * Palmer_Int * Davis5 + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longarmar)
#plot(mld_Longarmar)
#pp_check(mld_Longarmar, ndraws = 100)
saveRDS(mld_Longarmar, file = "Models/mld_Longarmar.rds")

#gaussian processing
##note, gaussian processing allows you to model multiple factors as a "surface"; so, you could do gp(lat, long), or gp(lat, long, elev) for a 3 dimensional space. this takes an immensely long time to process, so we will not run it here.

loo(mld_Longarma, mld_Longarmas, mld_Longarmar, compare = TRUE)
# Results:
#                  elpd_diff se_diff
# mld_Longarmas     0.0       0.0   
# mld_Longarmar    -0.8       1.8
# mld_Longarma     -4.1       3.3
# Differences within 2 SE; keep simpler mld_Longarma for ease of processing and interpreting

##let's see if variance and flood frequency are influenced by SST
#best linear model plus variance tested
#note the bf() part, this is used if there are multiple parts in the model
mld_Longarma_sig <- brm(bf(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
    # variance (sigma) model: does variability depend on SST?
    sigma ~ Seki_Int + Palmer_Int + Davis5),
  data   = clim,
  family = gaussian(),
  prior = c(
    prior(normal(63, 5), class = "Intercept"),
    prior(normal(0, 5), class = "b"),
    prior(normal(0, 1), class = "b", dpar = "sigma"),
    prior(student_t(3, 0, 1), class = "Intercept", dpar = "sigma")),
  chains = 4,
  iter   = 5000,
  warmup = 2500,
  seed = 2025,
  control = list(adapt_delta = 0.95))

saveRDS(mld_Longarma_sig, file = "Models/mld_Longarma_sig.rds")
readRDS("Models/mld_Longarma_sig.rds") -> mld_Longarma_sig
summary(mld_Longarma_sig)
# remember that sigma summaries default to log.

# Skew-normal model: check if lake elevation distribution is asymmetric.
# Note: skew-normal + ARMA was too heavy
mld_Long_skew <- brm(bf(elev ~ Seki_Int + Palmer_Int + Davis5,
    # right tail (alpha) model: do high elevations depend on SST?
    alpha ~ Seki_Int + Palmer_Int + Davis5),
  data   = clim,
  family = skew_normal(),
  chains = 4,
  iter   = 3000,
  warmup = 1000,
  seed = 2025)

saveRDS(mld_Long_skew, file = "Models/mld_Long_skew.rds")
readRDS("Models/mld_Long_skew.rds") -> mld_Long_skew
summary(mld_Long_skew)

##neither of these provide very strong results. Likely because ar/ma soaks up most of the year over year fluctuations; without year over year SSTs we don't see SST influence on skew and variability.

```

## Blind Prediction of Training Period
```{r blind_training_pred}
## Load the trained model, then predict the training period with elev set to NA. This shows what the model produces when it cannot observe elev values for ARMA, so similar to what we will do later for the hindcast, except that in this case we will be able to compare the prediction to observed values to see how accurate it is.

## ARMA prediction: what do arma terms in training contribute when predicting new data? When predicting out-of-sample, brms starts ARMA at zero, so none of the actual observations matter beyond the training window. Each draw simulates its own ARMA trajectory from there. Because the simulated arma terms are zero-mean, ARMA contributes nothing to the MEAN of fitted(), so the predicted mean is purely SST-driven. ARMA does influence CIs when using predict(), since each draw simulates noise based on what it learned from the arma terms. Since ours are ar circa .8 and ma -.1, the noise stabilizes and doesn't make the CIs too extremely wide. So, ultimately, arma does not inform the predicted mean, but it does inform prediction of the noise around the mean. 
# Note, this actually is handled in R by brms, not in Stan

mld_Longarma <- readRDS("Models/mld_Longarma.rds")

clim_train_blind <- clim %>%
  filter(!is.na(elev)) %>%
  arrange(Year_B2005) %>%
  mutate(elev_obs = elev, elev = NA)

blind_preds          <- as.data.frame(predict(mld_Longarma, newdata = clim_train_blind))
blind_preds$year_bp  <- clim_train_blind$Year_B2005
blind_preds$elev_obs <- clim_train_blind$elev_obs

ggplot(blind_preds, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.4, alpha = 0.4) +
  geom_smooth(aes(y = elev_obs), method = "loess", span = 500 / nrow(blind_preds),
              color = col_observed, linewidth = 0.9, se = FALSE) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.8) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Full Model: Blind Prediction of Training Period (elev = NA)",
       subtitle = "SST-driven mean only; ARMA initialized at zero (same as hindcast)") +
  theme_paashi()

```

## Exploring Periodicity (some assistance from Claude Code)
```{r exploring_periodicity}
## There is still a lot of noise in lake elevations (represented in our CIs), even when predicting from our current SST sets and with ARMA. Arma helps with understanding persistence of lake levels, but does not EXPLAIN deviations from the mean. When we actually look at a plot of lake elevation and listen to oral traditions, we know that the lake fluctuates year to year and has decadal and "generational" patterns. We should examine these patterns to see if we can explain shorter term variation / noise before we model deep into the past.

##Visual check: plot observed lake elevations with a line representing flood level; note ~decadal fluctuations and "generational" flood events.
clim <- clim[order(clim$Year_B2005), ] #make double sure clim is read in order of year

drainage_level <- 64.5
obs_elev <- clim %>% filter(!is.na(elev))

ggplot(obs_elev, aes(x = Calendar_Year)) +
  geom_line(aes(y = elev), color = "gray70", linewidth = 0.3) +
  geom_line(aes(y = pmax(elev, drainage_level)),
            color = col_observed, linewidth = 0.4) +
  geom_hline(yintercept = drainage_level, color = col_observed,
             linewidth = 0.3, linetype = "solid") +
  annotate("text", x = max(obs_elev$Calendar_Year) + 5, y = drainage_level + 0.3,
           label = "64.5 MaSL — Drainage Level",
           size = 2.5, hjust = 1, color = "gray40") +
  labs(x = "Calendar Year AD", y = "Water Surface Elevation (MaSL)",
       title = "Observed Tulare Lake Elevation — Drainage Periods Highlighted") +
  theme_paashi()


##check longer acf and pacf to see if multi-decade patterns exist
y <- clim$elev
y_complete <- stats::na.omit(y) #acf and pacf will not work if there are NAs

acf(y_complete, lag.max = 100, main = "century ACF of elev")
pacf(y_complete, lag.max = 100, main = "century PACF of elev")

#check periodogram to see if there are strong frequencies/periodicities in the data
spec.pgram(y_complete, log = "no", main = "Periodogram")

# Smoothed periodogram plotted with x-axis in years. Focus on periods >= 10 years where the interesting structure is
sp <- spec.pgram(y_complete, spans = c(7, 7), plot = FALSE)
sp_period <- 1 / sp$freq
sp_mask <- sp_period >= 10 & sp_period <= 200
plot(sp_period[sp_mask], sp$spec[sp_mask], type = "l",
     xlab = "Period (years)", ylab = "Spectrum",
     main = "Smoothed Periodogram (10–200 year periods)")
## Two clear peaks at ~35-40 years and ~70-80 years. The 70-80 year peak is likely the harmonic (double) of the ~40-year cycle. Rise at 150+ is century-scale variance, probably SST-driven.

# AR-based spectrum for comparison
sp_ar <- spec.ar(y_complete, plot = FALSE)
sp_ar_period <- 1 / sp_ar$freq
sp_ar_mask <- sp_ar_period >= 10 & sp_ar_period <= 200
plot(sp_ar_period[sp_ar_mask], sp_ar$spec[sp_ar_mask], type = "l",
     xlab = "Period (years)", ylab = "Spectrum",
     main = "AR-fitted Spectral Density (10–200 year periods)")
# Smooth curve with no peaks. So, the peaks in the other periodogram are NOT explained by AR(1) persistence, suggesting periodicity beyond what our ARMA(1,1) model captures.

##Now let's assess the flood data quantitatively:

# Flood recurrence: how often does elev exceed 64.5m? min_gap_years: floods separated by fewer than this many years are grouped as one event to account for 1-2 year dips during a wet period
flood_threshold <- 64.5
min_gap_years <- 5
flood_years <- which(y_complete > flood_threshold)
flood_starts <- flood_years[c(TRUE, diff(flood_years) > min_gap_years)]
flood_gaps <- diff(flood_starts)
cat("Number of flood events (min gap =", min_gap_years, "years):", length(flood_starts), "\n")
cat("Mean return interval:", round(mean(flood_gaps), 1), "years\n")
cat("Median return interval:", round(median(flood_gaps), 1), "years\n")
cat("Range:", min(flood_gaps), "-", max(flood_gaps), "years\n")
hist(flood_gaps, breaks = 15, main = paste0("Flood Return Intervals (elev > ", flood_threshold, "m, min gap = ", min_gap_years, "yr)"),
     xlab = "Years between flood events", col = "skyblue")
# Result: 51 flood events over ~2,000 years. Mean return = 39.4 years, median = 32 years. Mode around 20-30 years. Right-skewed: most floods are generational (~25-40 years apart) but some gaps stretch to 80-120 years. Consistent with oral traditions of "once a generation" flooding. The right skew may reflect prolonged drought periods.

# Next: check if the ~40-year periodicity survives after removing SST forcing and AR(1)
# persistence. If it does, it's genuine cyclicity in the lake system itself.
mld_Longarma <- readRDS("Models/mld_Longarma.rds")

# Residual periodograms: by examining the periodicity of residuals, we are looking at the periodicity of effects NOT explained by the models. Peaks that survive in residuals represent  structure the model does not capture

arma_resids <- residuals(mld_Longarma, type = "ordinary")[, "Estimate"]

sp_resid <- spec.pgram(arma_resids, spans = c(7, 7), plot = FALSE)
sp_resid_period <- 1 / sp_resid$freq
sp_resid_mask <- sp_resid_period >= 10 & sp_resid_period <= 200
plot(sp_resid_period[sp_resid_mask], sp_resid$spec[sp_resid_mask], type = "l",
     xlab = "Period (years)", ylab = "Spectrum",
     main = "Residual Periodogram: SSTs + ARMA(1,1) Model")
# Result: ~35-40 year peak persists. neither SST forcing nor ARMA(1,1) persistence accounts for this cycle. Sharp peaks at ~10, 15, 20 years also emerge once ARMA strips away persistence.

# Also check residuals from the non-ARMA model for comparison
mld_Long <- readRDS("Models/mld_Long.rds")
long_resids <- residuals(mld_Long, type = "ordinary")[, "Estimate"]

sp_long_resid <- spec.pgram(long_resids, spans = c(7, 7), plot = FALSE)
sp_long_period <- 1 / sp_long_resid$freq
sp_long_mask <- sp_long_period >= 10 & sp_long_period <= 200
plot(sp_long_period[sp_long_mask], sp_long_resid$spec[sp_long_mask], type = "l",
     xlab = "Period (years)", ylab = "Spectrum",
     main = "Residual Periodogram: SSTs Only Model")
# Result: similar peaks at ~35-40 and ~70-80 years, but overall power is ~20x higher because autocorrelation remains in these residuals. The short-period spikes (10-20 yr) are present but small because arma is not accounting for persistence.

## Wavelet analysis — shows how periodicity changes over time
# install.packages("WaveletComp") # if not installed
library(WaveletComp)

# Build a data frame for WaveletComp (needs a column named for the series)
wavelet_df <- data.frame(elev = y_complete)
wt <- analyze.wavelet(wavelet_df, "elev",
                       loess.span = 0,    # no detrending (we want to see all structure)
                       dt = 1,            # 1-year time step
                       dj = 1/20,         # frequency resolution
                       lowerPeriod = 8,   # minimum period to examine
                       upperPeriod = 128, # maximum period
                       make.pval = TRUE,  # significance testing
                       n.sim = 100)       # Monte Carlo simulations for significance

wt.image(wt, color.key = "quantile",
         main = "Wavelet Power Spectrum of Lake Elevation",
         periodlab = "Period (years)",
         timelab = "Year index (0 = most recent)")

# Average wavelet power across time — shows which periods are strongest overall
wt.avg(wt, main = "Average Wavelet Power by Period")

# Wavelet results: The heatmap (x = time, y = period, color = power) shows where the ~32-40 year cycle is active across the 2,000-year record. White contour lines mark statistically significant regions (p < 0.05 vs red noise). The average wavelet power plot confirms a significant peak around 32-40 years, consistent with periodogram and flood recurrence findings. The cycle is quasi-periodic it recurs on roughly that timescale but is not perfectly regular throughout the record.

```

## PDO/ENSO Testing
```{r pdo_enso_test}
# Our century-scale SSTs force long-term mean lake levels. ARMA(1,1) accounts for year-to-year persistence because the basin buffers swings in rainfall. But there are still year-over-year fluctuations and periodicity that our models do not explain.

# Based on contemporary climate science, PDO (Pacific Decadal Oscillation) and ENSO (El Nino-Southern Oscillation) likely drive those year-over-year fluctuations. They are the annually resolved versions of the same Pacific climate signals our century-scale proxies show
# Palmer_Int (~600yr resolution) = long-term mean of western equatorial Pacific, where ENSO operates; ~7 year cycles
# Seki_Int (~400yr resolution) = long-term mean of NE Pacific / California margin, where PDO operates; ~20-30 year cycles

# Let's add annually resolved PDO and ENSO records as predictors and check if they explain what we are seeing. These only cover ~1,000 years, so we cannot use them for model training or predictions.
# Sources: MacDonald & Case 2005 (PDO, 993-1996 CE), Li et al. 2011 (ENSO, 900-2002 CE)

# Filter to rows where all predictors + PDO + ENSO are available (~1,000 years)
clim %>%
  filter(!is.na(elev), !is.na(Seki_Int), !is.na(Palmer_Int),
         !is.na(Davis5), !is.na(PDO), !is.na(ENSO)) %>%
  arrange(Year_B2005) -> clim_test

cat("Diagnostic subset:", nrow(clim_test), "rows\n")
cat("Calendar year range:", max(clim_test$Calendar_Year), "to", min(clim_test$Calendar_Year), "CE\n")

# Fit two models on the same years for comparison
# Model 1: baseline (SSTs + ARMA only)
mld_test_longarma <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
    data = clim_test, family = gaussian(),
    prior = paashi_priors_arma,
    chains = 4, iter = 4000, warmup = 2000,
    seed = 2025,
    control = list(adapt_delta = 0.95))

# Model 2: SSTs + PDO + ENSO + ARMA
mld_test_pdo_enso <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + PDO + ENSO + arma(p = 1, q = 1, time = Year_B2005),
    data = clim_test, family = gaussian(),
    prior = paashi_priors_arma,
    chains = 4, iter = 4000, warmup = 2000,
    seed = 2025,
    control = list(adapt_delta = 0.95))

# LOO comparison
loo(mld_test_longarma, mld_test_pdo_enso)

# summaries
summary(mld_test_longarma)
summary(mld_test_pdo_enso)

# Blind prediction comparison: set elev = NA so ARMA starts from zero (same as hindcast)
clim_test_blind <- clim_test %>% mutate(elev_obs = elev, elev = NA)

blind_base <- as.data.frame(predict(mld_test_longarma, newdata = clim_test_blind))
blind_base$year_bp  <- clim_test_blind$Year_B2005
blind_base$elev_obs <- clim_test_blind$elev_obs

blind_pdo_enso <- as.data.frame(predict(mld_test_pdo_enso, newdata = clim_test_blind))
blind_pdo_enso$year_bp  <- clim_test_blind$Year_B2005
blind_pdo_enso$elev_obs <- clim_test_blind$elev_obs

ggplot(blind_base, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3, alpha = 0.4) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.5) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Blind Prediction: SSTs + ARMA(1,1) Only",
       subtitle = "Black = observed; blue = predicted (elev = NA)") +
  theme_paashi()

ggplot(blind_pdo_enso, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3, alpha = 0.4) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.5) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Blind Prediction: SSTs + PDO + ENSO + ARMA(1,1)",
       subtitle = "Black = observed; blue = predicted (elev = NA)") +
  theme_paashi()

## PDO is strongly related to lake elevation (+0.79m per unit PDO [0.68, 0.91]), ENSO adds +0.20m per unit [0.16, 0.25]. LOO confirms both improve the model by a lot (~122 elpd over just multi-century SSTs with arma). The blind prediction shows that adding annually-resolved PDO and ENSO does not change the predicted mean (both models produce the same SST-driven trend), but it does help explain much more of the year-over-year noise that our century-scale SSTs miss.

# These records only cover ~1,000 years, so they cannot be used for the 20,000-year hindcast. The year-over-year fluctuations they explain are represented in the CIs of the hindcast.

```

## Using Models to Predict Reality
```{r predictions}
## Now that we understand what our model can tell us and what our CIs represent, let's do an out of sample prediction using Longarma.
#there are several ways to do a posterior prediction. posterior_predict() runs a full simulation based on a trained model, with thousands of draws, outputting every result. This is usually used to compare a model to actual observed results. predict() runs posterior_predict, but then just outputs the mean and upper/lower 95; it is most useful if you just want to go straight to plotting results. posterior_epred ("expected prediction") runs a simulation based on a trained model, gets the mean, then makes predictions based on the mean. This is typically used to make posterior predictions on large sets of new data, since it is significantly lighter than running a full posterior_predict, given that posterior_epred() is just relying on the mean of an already trained model, whereas posterior_predict() or predict() would run thousands of predictions on all the new data, then return the mean.Lastly, fitted() runs posterior_epred() and then summarizes it, much as predict does for posterior_predict.

#First, let's use predict() to visualize how good the model is when arma has observations to work with. This will not have new data, it is just to compare the model against observations

as.data.frame(predict(mld_Longarma)) -> Longarma_Predict #outputs a data frame

#add time and observed data
Longarma_Predict$year_bp <- mld_Longarma$data$Year_B2005
Longarma_Predict$elev_obs <- mld_Longarma$data$elev

#create a csv that we can load later for plotting, rather than running this again
write.csv(Longarma_Predict, "Processed Data/Longarma_Predict.csv", row.names = FALSE)
#for later:
read.csv("Processed Data/Longarma_Predict.csv") -> Longarma_Predict

##Let's visualize the results quickly. We will do more in depth plotting later
#plot light blue ribbon of lower and upper 95, dark blue line is mean, and gray "highlight" is actual values
ggplot(Longarma_Predict, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = Estimate), color = "blue", linewidth = 1) +
  geom_line(aes(y = elev_obs), color = "black", linewidth = 1.5, alpha = 0.4) +
  labs(x = "Years Before Present", y = "Lake Elevation (MaSL)",title = "Observed vs. Modeled Lake Elevation") +
  theme_minimal()

##now let's run it back through time with "new data" from the rest of the climate dataset

#Build clean "new data" for prediction
newdata_pred <- clim %>%
  select(Year_B2005, Seki_Int, Palmer_Int, Davis5, elev) %>%
  filter(!is.na(Year_B2005), !is.na(Seki_Int), !is.na(Palmer_Int), !is.na(Davis5)) %>%
  arrange(Year_B2005)

# fitted() gives the expected mean trend with uncertainty about where the average is.
# predict() gives individual year prediction intervals with noise.

# Expected trend (mean uncertainty only)
Longarma_Fitted <- as.data.frame(fitted(mld_Longarma, newdata = newdata_pred))
colnames(Longarma_Fitted) <- c("Fit_Est", "Fit_SE", "Fit_Q2.5", "Fit_Q97.5")
Longarma_Fitted$year_bp <- newdata_pred$Year_B2005

# Full prediction intervals, uncertainty + variability
Longarma_FullPred <- as.data.frame(predict(mld_Longarma, newdata = newdata_pred))
colnames(Longarma_FullPred) <- c("Pred_Est", "Pred_SE", "Pred_Q2.5", "Pred_Q97.5")
Longarma_FullPred$year_bp <- newdata_pred$Year_B2005

# Combine into one data frame for plotting
hindcast <- left_join(Longarma_Fitted, Longarma_FullPred, by = "year_bp")
hindcast$elev_obs <- newdata_pred$elev

write.csv(hindcast, "Processed Data/Hindcast_Full.csv", row.names = FALSE)
hindcast <- read.csv("Processed Data/Hindcast_Full.csv")

# two-ribbon plot: outer = prediction interval, inner = credible interval for mean
ggplot(hindcast, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5),
              fill = col_model_ribbon, alpha = 0.3) +
  geom_ribbon(aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5),
              fill = col_model_inner, alpha = 0.5) +
  geom_line(aes(y = Fit_Est), color = col_model_line, linewidth = 0.5) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.5,
            alpha = 0.7, na.rm = TRUE) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Lake Elevation Hindcast: Expected Trend and Prediction Intervals",
       subtitle = "Inner band (dark) = 95% CI for mean; Outer band (light) = 95% prediction interval; Dark = observed") +
  theme_paashi()

```

## Publication Figures

### Full Hindcast Figure (assisted by Claude Code)
```{r fig4_hindcast, fig.width = 7.5, fig.height = 5}
hindcast <- read.csv("Processed Data/Hindcast_Full.csv")
training_boundary <- 2006
pleis_holo        <- 11700

# Panel A: Holocene (0-12,000 yr BP)
fig4a <- ggplot(hindcast %>% filter(year_bp <= 12000), aes(x = year_bp)) +
  annotate("rect", xmin = 0, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5),
              fill = col_model_ribbon, alpha = 0.4) +
  geom_ribbon(aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5),
              fill = col_model_inner, alpha = 0.5) +
  geom_line(aes(y = Fit_Est), color = col_model_line, linewidth = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3,
            alpha = 0.7, na.rm = TRUE) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.3) +
  annotate("text", x = training_boundary + 200, y = 69,
           label = "Training\nboundary", size = 2.2, hjust = 0, color = "gray40") +
  annotate("text", x = pleis_holo - 200, y = 69,
           label = "Pleistocene-\nHolocene\nboundary", size = 2.2,
           hjust = 1, color = col_pleis_line) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 12000, by = 2000)) +
  coord_cartesian(ylim = c(50, 72)) +
  labs(x = NULL, y = "Lake Elevation (m asl)",
       title = "A. Holocene (0\u201312,000 yr BP)") +
  theme_paashi()

# Panel B: Full range (0-23,000 yr BP)
fig4b <- ggplot(hindcast, aes(x = year_bp)) +
  annotate("rect", xmin = 0, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5),
              fill = col_model_ribbon, alpha = 0.4) +
  geom_ribbon(aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5),
              fill = col_model_inner, alpha = 0.5) +
  geom_line(aes(y = Fit_Est), color = col_model_line, linewidth = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3,
            alpha = 0.7, na.rm = TRUE) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.3) +
  annotate("text", x = 18000, y = 70,
           label = "Model trained on Holocene data;\ndeeper predictions carry\ngreater uncertainty",
           size = 2.0, color = "gray40", fontface = "italic") +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 22000, by = 4000)) +
  coord_cartesian(ylim = c(50, 72)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Lake Elevation (m asl)",
       title = "B. Full reconstruction (0\u201323,000 yr BP)") +
  theme_paashi()

fig4 <- fig4a / fig4b +
  plot_annotation(
    title = "Bayesian Hindcast of Tulare Lake Elevation",
    subtitle = paste0("Dark ribbon = 95% CI for expected trend (fitted);",
                      "  Light ribbon = 95% prediction interval;",
                      "  Dark line = observed data"))
fig4
## save_fig(fig4, "Figure_4_Hindcast.tiff", width_mm = 190, height_mm = 160)
```

### Archaeological Reconstruction Comparison (Assisted by Claude Code)
```{r fig5_arch_comparison, fig.width = 7.5, fig.height = 3.5}
combo <- read.csv("Processed Data/Model_Outputs_Combined.csv")
arch  <- read.csv("Processed Data/Lake12All.csv")
# hindcast has full prediction intervals; combo only has them for the training period
hindcast_holo <- hindcast %>% filter(year_bp <= 12000)
combo_holo <- combo %>% filter(year_b2005 <= 12000)
arch_holo  <- arch %>% filter(ybp <= 12000)

fig5 <- ggplot() +
  annotate("rect", xmin = 0, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  # Archaeological raw evidence
  geom_point(data = arch_holo, aes(x = ybp, y = elev),
             color = "gray75", size = 0.1, alpha = 0.3) +
  # Full prediction interval from hindcast data
  geom_ribbon(data = hindcast_holo,
              aes(x = year_bp, ymin = Pred_Q2.5, ymax = Pred_Q97.5),
              fill = col_model_ribbon, alpha = 0.5) +
  # Model trend
  geom_line(data = combo_holo,
            aes(x = year_b2005, y = Fitted_Estimate),
            color = col_model_line, linewidth = 0.4, alpha = 0.8) +
  # Archaeological reconstruction
  geom_line(data = combo_holo %>% filter(!is.na(arch_model)),
            aes(x = year_b2005, y = arch_model),
            color = col_arch, linewidth = 0.6, alpha = 0.85) +
  # Observed lake elevation
  geom_line(data = combo_holo %>% filter(!is.na(elev_obs)),
            aes(x = year_b2005, y = elev_obs),
            color = col_observed, linewidth = 0.25, alpha = 0.5) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.3) +
  annotate("text", x = pleis_holo - 200, y = 71,
           label = "Pleistocene-\nHolocene boundary",
           size = 2.0, hjust = 1, color = col_pleis_line) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 12000, by = 2000)) +
  coord_cartesian(ylim = c(48, 73)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Lake Elevation (m asl)",
       title = "Model Reconstruction vs. Archaeological Evidence",
       subtitle = paste0("Blue ribbon = 95% prediction interval;  ",
                         "Blue line = expected trend;  ",
                         "Terra cotta = archaeological reconstruction (Negrini et al.);  ",
                         "Gray = observed")) +
  theme_paashi()
fig5
## save_fig(fig5, "Figure_5_Arch_Comparison.tiff", width_mm = 190, height_mm = 100)
```

## Session Info
```{r session_info}
sessionInfo()
```