---
title: "Paashi_Paleo_Models_Pub"
author: "Robin_ML"
date: "2025-12-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages and Raw Data
```{r loading packages/data}

library(tidyverse)
library(tidybayes)
library(ggplot2)
library(ggthemes)
library(brms)
library(bayesplot)
library(loo)
library(patchwork)
library(scales)
library(ggrepel)
library(maps)

clim <- read.csv("Raw Data/ElevSSTs.csv")

```

## Publication Theme and Palette (assisted by Claude Code)
```{r theme_and_palette}
# Colorblind-safe, grayscale-distinguishable palette
col_model_line   <- "#1F4E79"   # dark blue — model trend line
col_model_ribbon <- "#A8C8E8"   # light blue — prediction interval
col_model_inner  <- "#4682B4"   # steel blue — fitted CI
col_observed     <- "gray20"    # near-black — observed data
col_arch         <- "#B85C38"   # terra cotta — archaeological reconstruction
col_seki         <- "#D4930D"   # amber — Seki SST (California Margin)
col_palmer       <- "#1B998B"   # teal — Palmer SST (W. Equatorial Pacific)
col_davis5       <- "#5B7B9A"   # slate blue — Davis5 SST (Sea of Okhotsk)
col_train_bg     <- "#FFF8F0"   # faint warm — training window background
col_pleis_line   <- "gray50"    # Pleistocene-Holocene boundary
col_pdo_line     <- "#B85C38"
col_wb_draw      <- "#4682B4"

# Shared theme for JAS (Elsevier) publication — min 7pt text, sans-serif
theme_paashi <- function(base_size = 8) {
  theme_minimal(base_size = base_size, base_family = "sans") %+replace%
    theme(
      plot.title       = element_text(size = rel(1.15), face = "bold",
                                      hjust = 0, margin = margin(b = 4)),
      plot.subtitle    = element_text(size = rel(0.95), color = "gray30",
                                      hjust = 0, margin = margin(b = 6)),
      axis.title       = element_text(size = rel(1.0)),
      axis.text        = element_text(size = rel(0.9), color = "gray20"),
      strip.text       = element_text(size = rel(1.0), face = "bold"),
      legend.text      = element_text(size = rel(0.9)),
      legend.title     = element_text(size = rel(0.95), face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
      plot.margin      = margin(8, 8, 8, 8),
      legend.position  = "bottom",
      legend.key.size  = unit(0.4, "cm")
    )
}

# Save figures at JAS dimensions
save_fig <- function(plot, filename,
                     width_mm = 190, height_mm = 120,
                     dpi = 500, device = "tiff") {
  ggsave(
    filename = file.path("Figures", filename),
    plot     = plot,
    width    = width_mm, height = height_mm,
    units    = "mm", dpi = dpi, device = device,
    compression = if (device == "tiff") "lzw" else NULL
  )
}

dir.create("Figures", showWarnings = FALSE)
```

## Checking Paleoclimate Data
```{r plotting}

par(mfrow = c(3, 2))
plot(clim$Year_B2005, clim$Seki_Int, main = "Seki (NE Pacific, 28kya, ~400yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Palmer_Int, main = "Palmer (W Eq Pacific, 23kya, ~600yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Davis1, main = "Davis-1/Barron (NE Pacific, 15kya, ~100yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Davis5, main = "Davis-5 (Okhotsk, 25kya, ~300yr res)", pch = ".", xlim = c(0, 25000))
plot(clim$Year_B2005, clim$Isono_Int, main = "Isono (Japan, 15kya, ~20yr res) - EXCLUDED", pch = ".", xlim = c(0, 25000))
par(mfrow = c(1, 1))


##Examine training data
fig_elev_hist <- ggplot(clim, aes(x = elev)) +
  geom_histogram(bins = 300, fill = col_model_ribbon, color = "white", linewidth = 0.1) +
  labs(x = "Lake Elevation (m asl)", y = "Count",
       title = "Distribution of Observed Lake Elevation (Training Period)") +
  theme_paashi()
fig_elev_hist
save_fig(fig_elev_hist, "Figure_S_Elevation_Histogram.tiff", height_mm = 90)
#This gives a normal distribution, so no further action needed; we can use family = gaussian. The distribution may in fact be bimodal with two very close peaks near the center-- this may represent "drought" and "flood" periods, as truly average rainfall is rarity in California according to oral traditions. However, these peaks are close enough to treat as a single normal distribution; bimodal modeling produces bad results

```

## Define Explicit Priors
```{r define_priors}
## Inspect what brms would choose as defaults for our final model
default_priors <- get_prior(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
  data = clim,
  family = gaussian())

print(default_priors)

## Define weakly informative priors in case others have different versions of R or BRMS
# Intercept: Normal(63, 5) -- centered on mean observed lake elevation (~63 MaSL), 95% prior mass between 53-73m, generous given basin physical limits. Modern basin bottom is at 55masl, and highest visible geological shoreline is ~70masl. Note that over time, lake basins may rise due to sediment.
# Slopes: Normal(0, 5) -- 1-degree SST change shifts lake by <10m (2 SD);  generous since entire observed elev range is ~9m
# Sigma: half-Student-t(3, 0, 3) -- standard weakly informative for residual SD
# AR/MA: brms defaults to uniform(-1, 1)

# Base priors for models without autocorrelation terms
paashi_priors <- c(
  prior(normal(63, 5), class = "Intercept"),
  prior(normal(0, 5), class = "b"),
  prior(student_t(3, 0, 3), class = "sigma"))

# Extended priors for ARMA models (ar/ma must be explicit for prior predictive checks)
paashi_priors_arma <- c(
  paashi_priors,
  prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1),
  prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1))

```

## Prior Predictive Check
```{r prior_predictive_check}
# Fit model with sample_prior = "only" to draw from priors without seeing data to verify our priors are weakly informative but plausible
mld_prior_only <- brm(
  elev ~ Seki_Int + Palmer_Int + Davis5 +
    arma(p = 1, q = 1, time = Year_B2005),
  data = clim,
  family = gaussian(),
  prior = paashi_priors_arma,
  sample_prior = "only",
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 2025)

# Simulated data should include the observed range but not be too close, as these priors are weakly informative
fig_prior_pp <- pp_check(mld_prior_only, ndraws = 100) +
  labs(title = "Prior Predictive Check",
       subtitle = "Dark = observed; light = simulated from prior only") +
  xlim(30, 100) +
  theme_paashi()
fig_prior_pp
save_fig(fig_prior_pp, "Figure_S_Prior_Predictive.tiff", height_mm = 90)
## looks good!

```

## Create and Test Models
```{r model testing}
### running brms on paleoclimate datasets
##model sea surface temperature relationships to elevation


##Single-Predictor Models
# Test each SST record individually

# Seki (NE Pacific / California Margin, longest record)
mld_Seki <- brm(elev ~ Seki_Int,
                data = clim,
                family = gaussian(),
                prior = paashi_priors,
                chains = 4, iter = 3000, warmup = 1000,
                seed = 2025)
summary(mld_Seki)

# Palmer (Western Equatorial Pacific, ENSO-related signal)
mld_Palmer <- brm(elev ~ Palmer_Int,
                  data = clim,
                  family = gaussian(),
                  prior = paashi_priors,
                  chains = 4, iter = 3000, warmup = 1000,
                  seed = 2025)
summary(mld_Palmer)

# Davis1/Barron (NE Pacific, highest resolution among usable records)
mld_Davis1 <- brm(elev ~ Davis1,
                  data = clim,
                  family = gaussian(),
                  prior = paashi_priors,
                  chains = 4, iter = 3000, warmup = 1000,
                  seed = 2025)
summary(mld_Davis1)

# Davis5 (Sea of Okhotsk, long record with good resolution)
mld_Davis5 <- brm(elev ~ Davis5,
                  data = clim,
                  family = gaussian(),
                  prior = paashi_priors,
                  chains = 4, iter = 3000, warmup = 1000,
                  seed = 2025)
summary(mld_Davis5)

## Compare Single-Predictor Models
loo(mld_Seki, mld_Palmer, mld_Davis1, mld_Davis5, compare = TRUE)
#Model comparisons:
#           elpd_diff se_diff
#mld_Davis5   0.0       0.0  
#mld_Davis1  -1.9       7.5  
#mld_Palmer  -5.8       5.8  
#mld_Seki   -21.0       6.2


## Combined-Predictor Models

# mld_Long: Three longest records (for extended hindcast to ~23kya)
# Seki (28kya) + Palmer (23kya) + Davis5 (25kya)
mld_Long <- brm(elev ~ Seki_Int + Palmer_Int + Davis5,
                data = clim,
                family = gaussian(),
                prior = paashi_priors,
                chains = 4, iter = 3000, warmup = 1000,
                seed = 2025)
summary(mld_Long)
saveRDS(mld_Long, file = "Models/mld_Long.rds")

# mld_Short: All four records (best fit, limited to ~15kya by Davis1)
# Adds Davis1/Barron for improved fit within Holocene
mld_Short <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + Davis1,
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors,
                 chains = 4, iter = 3000, warmup = 1000,
                 seed = 2025)
summary(mld_Short)


#model 2 closest to California: Seki (NE Pacific "California Margin"), Davis1/Barron (California Margin)
mld_Cal <- brm(elev ~ Seki_Int + Davis1,
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors,
                 chains = 4,
                 iter = 3000,
                 warmup = 1000,
                 seed = 2025)
summary(mld_Cal)


## Compare Combined Models
loo(mld_Long, mld_Short, mld_Cal, compare = TRUE)
#Model comparisons:
#          elpd_diff se_diff
#mld_Short   0.0       0.0  
#mld_Long  -15.8       5.6  
#mld_Cal   -26.1       7.3

## Based off of these results, mld_Long looks like the way to go. It is not too terrible compared to mld_Short, and will allow us to predict much further back in time.

```

### Figure 1: Study Area and Data Overview (assisted by Claude Code)
```{r fig1_map_and_sst, fig.width = 7.5, fig.height = 6}

##Let's visualize where our selected SSTs are coming from and what they look like.

# Panel A: Pacific Basin map with SST core sites + Tulare Lake
sites <- data.frame(
  name = c("Tulare Lake", "Seki et al.\n(California Margin)",
           "Palmer & Pearson\n(W. Equatorial Pacific)",
           "Davis et al.\n(Sea of Okhotsk)"),
  lon  = c(-119.5, -126.0, 159.0, 145.0),
  lat  = c(35.9, 42.0, 0.0, 50.0),
  type = c("lake", "sst", "sst", "sst"),
  col  = c(col_observed, col_seki, col_palmer, col_davis5)
)

world2 <- map_data("world2")
sites$lon360 <- ifelse(sites$lon < 0, sites$lon + 360, sites$lon)

fig1a <- ggplot() +
  geom_polygon(data = world2, aes(x = long, y = lat, group = group),
               fill = "gray92", color = "gray70", linewidth = 0.15) +
  geom_point(data = sites[1,], aes(x = lon360, y = lat),
             color = sites$col[1], shape = 18, size = 3.5) +
  geom_point(data = sites[-1,], aes(x = lon360, y = lat),
             color = sites$col[-1], fill = NA,
             shape = 21, size = 4.5, stroke = 1.5) +
  geom_text_repel(data = sites, aes(x = lon360, y = lat, label = name),
                  size = 2.2, color = sites$col,
                  box.padding = 0.5, max.overlaps = 10,
                  fontface = "bold", lineheight = 0.85) +
  coord_cartesian(xlim = c(100, 300), ylim = c(-15, 58), expand = FALSE) +
  scale_x_continuous(breaks = c(120, 150, 180, 210, 240, 270),
                     labels = c("120\u00B0E", "150\u00B0E", "180\u00B0",
                                "150\u00B0W", "120\u00B0W", "90\u00B0W")) +
  labs(title = "A. Study area and paleoclimate data sources",
       x = NULL, y = NULL) +
  theme_paashi() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 6),
        panel.grid.major = element_line(color = "gray85", linewidth = 0.2))

# Panel B: SST predictor time series
sst_interp <- clim %>%
  select(Year_B2005, Seki_Int, Palmer_Int, Davis5) %>%
  pivot_longer(-Year_B2005, names_to = "source", values_to = "sst_interp") %>%
  mutate(source = recode(source,
    "Seki_Int"   = "Seki (California Margin)",
    "Palmer_Int" = "Palmer (W. Equatorial Pacific)",
    "Davis5"     = "Davis (Sea of Okhotsk)"))

sst_raw <- clim %>%
  select(Year_B2005, Seki_SST, Palmer_SST) %>%
  pivot_longer(-Year_B2005, names_to = "source", values_to = "sst_raw") %>%
  filter(!is.na(sst_raw)) %>%
  mutate(source = recode(source,
    "Seki_SST"  = "Seki (California Margin)",
    "Palmer_SST" = "Palmer (W. Equatorial Pacific)"))

sst_colors <- c(
  "Seki (California Margin)"       = col_seki,
  "Palmer (W. Equatorial Pacific)"  = col_palmer,
  "Davis (Sea of Okhotsk)"          = col_davis5)

fig1b <- ggplot() +
  geom_line(data = sst_interp,
            aes(x = Year_B2005, y = sst_interp, color = source),
            linewidth = 0.3, alpha = 0.7) +
  geom_point(data = sst_raw,
             aes(x = Year_B2005, y = sst_raw, color = source),
             size = 0.4, alpha = 0.5) +
  scale_color_manual(values = sst_colors) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 28000, by = 4000)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Sea Surface Temperature (\u00B0C)",
       title = "B. Pacific SST records used as predictors",
       subtitle = "Lines = interpolated; points = original measurements",
       color = NULL) +
  theme_paashi() +
  theme(legend.position = "bottom")

fig1 <- fig1a / fig1b + plot_layout(heights = c(3, 3))
fig1
save_fig(fig1, "Figure_1_Data_Overview.tiff", width_mm = 190, height_mm = 170)
```

##Add Prior Year Influence
```{r refining_models}
##Now that we have selected our SST sets with the best correlation to lake elevation, we want to see if other factors play in. Because this is a hydrological system, it is likely that the prior year can influence the next year.

##first, let's check our "auto correlation function" or ACF, to see how correlated each year's value is to the prior year; this will help us design better models.

#fix data for acf and pacf
clim <- clim[order(clim$Year_B2005), ] #make double sure clim is read in order of year
y <- clim$elev
y_complete <- stats::na.omit(y) #acf and pacf will not work if there are NAs

# acf checks auto correlation to see how long a value can influence subsequent values. PACF (partial ACF) removes the overlap so you can see more clearly the direct influence of prior years.
acf(y_complete, main = "ACF of elev (NAs dropped)")
pacf(y_complete, main = "PACF of elev (NAs dropped)")

#looks like we are strongly correlated for 1 year prior in the PACF. So, in ACF, value is strongly correlated with prior values, slowly tapering until ~15 prior values.
#periodicity around ~40 years backs up oral traditions, but is not necessarily strong enough to warrant adding fourier terms or seasonal arma. We will investigate periodicity in a bit.

## now we can run our best elev ~ SST model, adding influence from prior years. There are two arguments for adding in auto-correlation to a model. ar(p=x) says that the current value is correlated with past VALUES of x prior years. ma(q=x) says that the current value is correlated with past SHOCKS, or large CHANGES in value from the expected/mean value, from x prior years. arma(p = , q = ) just combines both. We can determine our p and q values from our ACF and PACF results.

#because we have a big spike in correlation for the previous year, plus a small correlation with the prior year in PACF; and a smooth taper in ACF, we can see that what matters most strongly is the VALUE of the prior year. If there were multiple spikes before taper in ACF, then we might pay more attention to our q value for ma().

#given what we see here, we can probably be safe with ar(p = 1), but could also reasonably try arma(p = 1, q = 1). 

##model Long with previous year effects
#ar(p = 1) for long, smooth persistence, good in hydrological models
mld_Longar <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + ar(p = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = c(paashi_priors,
                           prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1)),
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longar)
#plot(mld_Longar)
#pp_check(mld_Longar, ndraws = 100)


#ma(q = 1) does the deviation from the mean matter more than the actual value? Let's check ma
mld_Longma <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + ma(q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = c(paashi_priors,
                           prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1)),
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longma)
#plot(mld_Longma)
#pp_check(mld_Longma, ndraws = 100)

#arma(p = 1, q = 1) combines to see if both matter.
mld_Longarma <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longarma)
plot(mld_Longarma)
pp_check(mld_Longarma, ndraws = 100)
saveRDS(mld_Longarma, file = "Models/mld_Longarma.rds")
readRDS("Models/mld_Longarma.rds") -> mld_Longarma

loo(mld_Long, mld_Longar, mld_Longma, mld_Longarma, compare = TRUE)
# Results:
#              elpd_diff se_diff
# mld_Longarma    0.0       0.0 
# mld_Longar     -4.1       3.4 
# mld_Longma   -405.7      30.9 
# mld_Long     -998.7      42.5

##Longarma is best!

```

### Model Diagnostics Figure (assisted by Claude Code)
```{r figS1_diagnostics, fig.width = 7.5, fig.height = 8}
mld_Longarma <- readRDS("Models/mld_Longarma.rds")

# Panel A: Trace plots
s1a <- mcmc_trace(mld_Longarma,
                  pars = c("b_Intercept", "b_Seki_Int", "b_Palmer_Int",
                           "b_Davis5", "sigma", "ar[1]", "ma[1]"),
                  facet_args = list(ncol = 1)) +
  labs(title = "A. Trace plots") +
  theme_paashi(base_size = 7)

# Panel B: Posterior predictive density overlay
s1b <- pp_check(mld_Longarma, ndraws = 100, type = "dens_overlay") +
  labs(title = "B. Posterior predictive check") +
  theme_paashi(base_size = 7)

# Panel C: ACF posterior predictive check
yrep   <- posterior_predict(mld_Longarma, ndraws = 200)
y_obs  <- mld_Longarma$data$elev
acf_obs_vals <- acf(y_obs[!is.na(y_obs)], lag.max = 20, plot = FALSE)$acf[,,1]

acf_reps <- apply(yrep[1:200, !is.na(y_obs)], 1, function(y) {
  acf(y, lag.max = 20, plot = FALSE)$acf[,,1]
})

acf_rep_df <- as.data.frame(t(acf_reps)) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "lag", values_to = "acf") %>%
  mutate(lag = as.integer(gsub("V", "", lag)) - 1)

acf_obs_df <- data.frame(lag = 0:20, acf = acf_obs_vals)

s1c <- ggplot() +
  geom_line(data = acf_rep_df, aes(x = lag, y = acf, group = draw),
            alpha = 0.08, color = "gray50") +
  geom_line(data = acf_obs_df, aes(x = lag, y = acf),
            color = "black", linewidth = 0.8) +
  labs(x = "Lag (years)", y = "Autocorrelation",
       title = "C. Posterior predictive ACF check",
       subtitle = "Black = observed; gray = simulated") +
  theme_paashi(base_size = 7)

# Panel D: LOO Pareto k diagnostics
loo_longarma <- loo(mld_Longarma)
k_values <- loo_longarma$diagnostics$pareto_k
k_df <- data.frame(obs = seq_along(k_values), k = k_values)

s1d <- ggplot(k_df, aes(x = obs, y = k)) +
  geom_point(size = 0.3, alpha = 0.4, color = col_model_inner) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "orange",
             linewidth = 0.4) +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "red",
             linewidth = 0.4) +
  annotate("text", x = nrow(k_df) * 0.85, y = 0.55,
           label = "k = 0.5 (caution)", size = 2.2, color = "orange") +
  annotate("text", x = nrow(k_df) * 0.85, y = 0.75,
           label = "k = 0.7 (unreliable)", size = 2.2, color = "red") +
  labs(x = "Observation", y = "Pareto k",
       title = "D. LOO-CV Pareto k diagnostics",
       subtitle = paste0("All ", nrow(k_df), " observations below 0.5")) +
  theme_paashi(base_size = 7)

figS1 <- (s1a | (s1b / s1c / s1d)) +
  plot_layout(widths = c(1, 1)) +
  plot_annotation(
    title = "Supplementary Figure S1: Model Diagnostics (mld_Longarma)")
figS1
save_fig(figS1, "Figure_S1_Diagnostics.tiff", width_mm = 190, height_mm = 220)
```

## Continue Model Testing
```{r model_testing_continued}
# Test whether SSTs are necessary, or if ARMA alone suffices
# No "b" class here since there are no slope predictors (intercept-only)
mld_arma_only <- brm(elev ~ 1 + arma(p = 1, q = 1, time = Year_B2005),
                     data = clim,
                     family = gaussian(),
                     prior = c(
                       prior(normal(63, 5), class = "Intercept"),
                       prior(student_t(3, 0, 3), class = "sigma"),
                       prior(uniform(-1, 1), class = "ar", lb = -1, ub = 1),
                       prior(uniform(-1, 1), class = "ma", lb = -1, ub = 1)),
                     chains = 4, iter = 4000, warmup = 2000,
                     seed = 2025,
                     control = list(adapt_delta = 0.95))

loo(mld_Longarma, mld_arma_only, compare = TRUE)
# Results: elpd_diff = -0.1 (SE 2.5) -- essentially identical for year-to-year prediction. This is expected in loo since ARMA dominates year over year correlation; loo (leave one out) observes prediction when one observation is left out at a time. SSTs are essential for century+ scale climate forcing, because arma alone does not inform the mean when making predictions on new data.

##let's check for non-linear and more complex relationships

#spline instead of linear
mld_Longarmas <- brm(elev ~ s(Seki_Int) + s(Palmer_Int) + s(Davis5) + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longarmas)
#plot(mld_Longarmas)
#pp_check(mld_Longarmas, ndraws = 100)
saveRDS(mld_Longarmas, file = "Models/mld_Longarmas.rds")

#linear interactions between variables
# Using * for full interaction hierarchy (main effects + 2-way + 3-way)
mld_Longarmar <- brm(elev ~ Seki_Int * Palmer_Int * Davis5 + arma(p = 1, q = 1, time = Year_B2005),
                 data = clim,
                 family = gaussian(),
                 prior = paashi_priors_arma,
                 chains = 4,
                 iter = 4000,
                 warmup = 2000,
                 seed = 2025,
                 control = list(adapt_delta = 0.95))

summary(mld_Longarmar)
#plot(mld_Longarmar)
#pp_check(mld_Longarmar, ndraws = 100)
saveRDS(mld_Longarmar, file = "Models/mld_Longarmar.rds")

#gaussian processing
##note, gaussian processing allows you to model multiple factors as a "surface"; so, you could do gp(lat, long), or gp(lat, long, elev) for a 3 dimensional space. this takes an immensely long time to process, so we will not run it here.

loo(mld_Longarma, mld_Longarmas, mld_Longarmar, compare = TRUE)
# Results:
#                  elpd_diff se_diff
# mld_Longarmas     0.0       0.0   
# mld_Longarmar    -0.8       1.8
# mld_Longarma     -4.1       3.3
# Differences within 2 SE; keep simpler mld_Longarma for ease of processing and interpreting

##let's see if variance and flood frequency are influenced by SST
#best linear model plus variance tested
#note the bf() part, this is used if there are multiple parts in the model
mld_Longarma_sig <- brm(bf(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
    # variance (sigma) model: does variability depend on SST?
    sigma ~ Seki_Int + Palmer_Int + Davis5),
  data   = clim,
  family = gaussian(),
  prior = c(
    prior(normal(63, 5), class = "Intercept"),
    prior(normal(0, 5), class = "b"),
    prior(normal(0, 1), class = "b", dpar = "sigma"),
    prior(student_t(3, 0, 1), class = "Intercept", dpar = "sigma")),
  chains = 4,
  iter   = 5000,
  warmup = 2500,
  seed = 2025,
  control = list(adapt_delta = 0.95))

saveRDS(mld_Longarma_sig, file = "Models/mld_Longarma_sig.rds")
readRDS("Models/mld_Longarma_sig.rds") -> mld_Longarma_sig
summary(mld_Longarma_sig)
# remember that sigma summaries default to log.

# Skew-normal model: check if lake elevation distribution is asymmetric.
# Note: skew-normal + ARMA was too heavy
mld_Long_skew <- brm(bf(elev ~ Seki_Int + Palmer_Int + Davis5,
    # right tail (alpha) model: do high elevations depend on SST?
    alpha ~ Seki_Int + Palmer_Int + Davis5),
  data   = clim,
  family = skew_normal(),
  chains = 4,
  iter   = 3000,
  warmup = 1000,
  seed = 2025)

saveRDS(mld_Long_skew, file = "Models/mld_Long_skew.rds")
readRDS("Models/mld_Long_skew.rds") -> mld_Long_skew
summary(mld_Long_skew)

##neither of these provide very strong results. Likely because ar/ma soaks up most of the year over year fluctuations; without year over year SSTs we don't see SST influence on skew and variability.

```

## Blind Prediction of Training Period
```{r blind_training_pred}
## Load the trained model, then predict the training period with elev set to NA. This shows what the model produces when it cannot observe elev values for ARMA, so similar to what we will do later for the hindcast, except that in this case we will be able to compare the prediction to observed values to see how accurate it is.

## ARMA prediction: what do arma terms in training contribute when predicting new data? When predicting out-of-sample, brms starts ARMA at zero, so none of the actual observations matter beyond the training window. Each draw simulates its own ARMA trajectory from there. Because the simulated arma terms are zero-mean, ARMA contributes nothing to the MEAN of fitted(), so the predicted mean is purely SST-driven. ARMA does influence CIs when using predict(), since each draw simulates noise based on what it learned from the arma terms. Since ours are ar circa .8 and ma -.1, the noise stabilizes and doesn't make the CIs too extremely wide. So, ultimately, arma does not inform the predicted mean, but it does inform prediction of the noise around the mean. 
# Note, this actually is handled in R by brms, not in Stan

mld_Longarma <- readRDS("Models/mld_Longarma.rds")

clim_train_blind <- clim %>%
  filter(!is.na(elev)) %>%
  arrange(Year_B2005) %>%
  mutate(elev_obs = elev, elev = NA)

blind_preds          <- as.data.frame(predict(mld_Longarma, newdata = clim_train_blind))
blind_preds$year_bp  <- clim_train_blind$Year_B2005
blind_preds$elev_obs <- clim_train_blind$elev_obs

fig_blind_train <- ggplot(blind_preds, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.4, alpha = 0.4) +
  geom_smooth(aes(y = elev_obs), method = "loess", span = 500 / nrow(blind_preds),
              color = col_observed, linewidth = 0.9, se = FALSE) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.8) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Full Model: Blind Prediction of Training Period (elev = NA)",
       subtitle = "SST-driven mean only; ARMA initialized at zero (same as hindcast)") +
  theme_paashi()
fig_blind_train
save_fig(fig_blind_train, "Figure_S_Blind_Training.tiff")

```

## Exploring Periodicity- Persistence, Climate Forcing, Basin Physics; What do our CIs contain?
```{r exploring_periodicity}
## There is still a lot of noise in lake elevations (represented in our CIs), even when predicting from our current SST sets and with ARMA. Arma helps with understanding persistence of lake levels, but does not EXPLAIN deviations from the mean. When we actually look at a plot of lake elevation and listen to oral traditions, we know that the lake fluctuates year to year and has decadal and "generational" patterns. We should examine these patterns to see if we can explain shorter term variation / noise before we model deep into the past.

##Visual check: plot observed lake elevations with a line representing flood level; note ~decadal fluctuations and "generational" flood events.
clim <- clim[order(clim$Year_B2005), ] #make double sure clim is read in order of year

drainage_level <- 64.5
obs_elev <- clim %>% filter(!is.na(elev))

fig_drainage <- ggplot(obs_elev, aes(x = Calendar_Year)) +
  geom_line(aes(y = elev), color = "gray70", linewidth = 0.3) +
  geom_line(aes(y = pmax(elev, drainage_level)),
            color = col_observed, linewidth = 0.4) +
  #geom_hline(yintercept = drainage_level, color = col_observed,
             #linewidth = 0.3, linetype = "solid") +
  labs(x = "Calendar Year AD", y = "Water Surface Elevation (MaSL)",
       title = "Observed Tulare Lake Elevation — Drainage Periods Highlighted") +
  theme_paashi()
fig_drainage
save_fig(fig_drainage, "Figure_S_Elevation_Drainage.tiff", width_mm = 190, height_mm = 80)

# Smoothed periodogram of observed elevations
y <- clim$elev
y_complete <- stats::na.omit(y)

sp <- spec.pgram(y_complete, spans = c(7, 7), plot = FALSE)
sp_period <- 1 / sp$freq
sp_mask <- sp_period >= 10 & sp_period <= 200
plot(sp_period[sp_mask], sp$spec[sp_mask], type = "l",
     xlab = "Period (years)", ylab = "Spectrum",
     main = "Smoothed Periodogram (10–200 year periods)")
## Two clear peaks at ~35-40 years and ~70-80 years. The 70-80 year peak is likely the harmonic (double) of the ~40-year cycle. Rise at 150+ is century-scale variance, probably SST-driven.

##Now let's assess the flood data quantitatively:

# Flood recurrence: how often does elev exceed 64.5m? min_gap_years: floods separated by fewer than this many years are grouped as one event to account for 1-2 year dips during a wet period
flood_threshold <- 64.5
min_gap_years <- 5
flood_years <- which(y_complete > flood_threshold)
flood_starts <- flood_years[c(TRUE, diff(flood_years) > min_gap_years)]
flood_gaps <- diff(flood_starts)
cat("Number of flood events (min gap =", min_gap_years, "years):", length(flood_starts), "\n")
cat("Mean return interval:", round(mean(flood_gaps), 1), "years\n")
cat("Median return interval:", round(median(flood_gaps), 1), "years\n")
cat("Range:", min(flood_gaps), "-", max(flood_gaps), "years\n")
flood_gap_df <- data.frame(gap = flood_gaps)
flood_mean <- mean(flood_gaps)
flood_median <- median(flood_gaps)
n_floods <- length(flood_starts)

fig_flood <- ggplot(flood_gap_df, aes(x = gap)) +
  geom_histogram(bins = 15, fill = col_model_ribbon, color = "white") +
  geom_vline(xintercept = flood_mean, linetype = "dashed", color = col_model_line, linewidth = 0.5) +
  geom_vline(xintercept = flood_median, linetype = "dotted", color = col_arch, linewidth = 0.5) +
  annotate("text", x = flood_mean + 2, y = Inf, vjust = 2, hjust = 0, size = 2.5,
           label = paste0("Mean = ", round(flood_mean, 1), " yr"), color = col_model_line) +
  annotate("text", x = flood_median - 2, y = Inf, vjust = 3.5, hjust = 1, size = 2.5,
           label = paste0("Median = ", round(flood_median, 1), " yr"), color = col_arch) +
  annotate("text", x = Inf, y = Inf, vjust = 1.5, hjust = 1.1, size = 2.5, color = "gray30",
           label = paste0("n = ", n_floods, " events over ~2,000 yr\n",
                          "Range: ", min(flood_gaps), "-", max(flood_gaps), " yr")) +
  labs(x = "Years Between Flood Events", y = "Count",
       title = paste0("Flood Return Intervals (elev > ", flood_threshold, " m asl)")) +
  theme_paashi()
fig_flood
save_fig(fig_flood, "Figure_S_Flood_Return.tiff", height_mm = 90)
# Result: 51 flood events over ~2,000 years. Mean return = 39.4 years, median = 32 years. Mode around 20-30 years. Right-skewed: most floods are generational (~25-40 years apart) but some gaps stretch to 80-120 years. Consistent with oral traditions of "once a generation" flooding. The right skew may reflect prolonged drought periods.

```

## PDO/ENSO Diagnostic: What Do Our Hindcast CIs Represent?
```{r pdo_enso_test}
## Our century-scale SSTs force long-term mean lake levels. ARMA(1,1) accounts for year-to-year persistence because the basin buffers swings in rainfall. But there are still year-over-year fluctuations and ~35-40 year periodicity that our models do not explain.

# Based on contemporary climate science, PDO (Pacific Decadal Oscillation) and ENSO (El Nino-Southern Oscillation) likely drive those fluctuations. They are the annually resolved versions of the same Pacific climate signals our century-scale proxies show:
# Palmer_Int (~600yr resolution) = long-term mean of western equatorial Pacific, where ENSO operates; ~7 year cycles
# Seki_Int (~400yr resolution) = long-term mean of NE Pacific / California margin, where PDO operates; ~20-30 year cycles

# To quantify what our hindcast CIs represent, we fit three models on the ~1,000-year window where annually-resolved PDO and ENSO reconstructions overlap with our training data, then compare blind predictions (elev = NA, same as hindcast).
# Sources: MacDonald & Case 2005 (PDO, 993-1996 CE), Li et al. 2011 (ENSO, 900-2002 CE)

# Filter to rows where all predictors + PDO + ENSO are available (~1,000 years)
clim %>%
  filter(!is.na(elev), !is.na(Seki_Int), !is.na(Palmer_Int),
         !is.na(Davis5), !is.na(PDO), !is.na(ENSO)) %>%
  arrange(Year_B2005) -> clim_test

# Model 1: SSTs + ARMA baseline (same structure as mld_Longarma, fit on this subset)
if (file.exists("Models/mld_test_longarma.rds")) {
  mld_test_longarma <- readRDS("Models/mld_test_longarma.rds")
} else {
  mld_test_longarma <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + arma(p = 1, q = 1, time = Year_B2005),
      data = clim_test, family = gaussian(),
      prior = paashi_priors_arma,
      chains = 4, iter = 4000, warmup = 2000,
      seed = 2025,
      control = list(adapt_delta = 0.95))
  saveRDS(mld_test_longarma, "Models/mld_test_longarma.rds")
}

# Model 2: PDO + ENSO only (no SSTs) — can annually-resolved indices alone track the lake?
if (file.exists("Models/mld_test_pdo_enso_only.rds")) {
  mld_test_pdo_enso_only <- readRDS("Models/mld_test_pdo_enso_only.rds")
} else {
  mld_test_pdo_enso_only <- brm(elev ~ PDO + ENSO + arma(p = 1, q = 1, time = Year_B2005),
      data = clim_test, family = gaussian(),
      prior = paashi_priors_arma,
      chains = 4, iter = 4000, warmup = 2000,
      seed = 2025,
      control = list(adapt_delta = 0.95))
  saveRDS(mld_test_pdo_enso_only, "Models/mld_test_pdo_enso_only.rds")
}

# Model 3: SSTs + PDO + ENSO + ARMA (full model for LOO reference)
if (file.exists("Models/mld_test_pdo_enso.rds")) {
  mld_test_pdo_enso <- readRDS("Models/mld_test_pdo_enso.rds")
} else {
  mld_test_pdo_enso <- brm(elev ~ Seki_Int + Palmer_Int + Davis5 + PDO + ENSO + arma(p = 1, q = 1, time = Year_B2005),
      data = clim_test, family = gaussian(),
      prior = paashi_priors_arma,
      chains = 4, iter = 4000, warmup = 2000,
      seed = 2025,
      control = list(adapt_delta = 0.95))
  saveRDS(mld_test_pdo_enso, "Models/mld_test_pdo_enso.rds")
}

# LOO comparison
loo(mld_test_longarma, mld_test_pdo_enso_only, mld_test_pdo_enso)

# Coefficient comparison
summary(mld_test_longarma)
summary(mld_test_pdo_enso_only)
summary(mld_test_pdo_enso)
## Loo and summaries show that SSTs don't add anything to the PDO/ENSO predictions. This is to be expected, as the SSTs are essentially just a less detailed record of the same climate signals.

# Blind prediction comparison: set elev = NA so ARMA starts from zero, same as the hindcast.
clim_test_blind <- clim_test %>% mutate(elev_obs = elev, elev = NA)

blind_sst <- as.data.frame(predict(mld_test_longarma, newdata = clim_test_blind))
blind_sst$year_bp  <- clim_test_blind$Year_B2005
blind_sst$elev_obs <- clim_test_blind$elev_obs

blind_pdo <- as.data.frame(predict(mld_test_pdo_enso_only, newdata = clim_test_blind))
blind_pdo$year_bp  <- clim_test_blind$Year_B2005
blind_pdo$elev_obs <- clim_test_blind$elev_obs

## SST+ARMA vs PDO+ENSO+ARMA
p_blind_sst <- ggplot(blind_sst, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3, alpha = 0.4) +
  geom_line(aes(y = Estimate), color = col_model_line, linewidth = 0.5) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Blind Prediction: SSTs + ARMA(1,1)") +
  theme_paashi()

p_blind_pdo <- ggplot(blind_pdo, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = col_model_ribbon, alpha = 0.4) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.3, alpha = 0.4) +
  geom_line(aes(y = Estimate), color = col_pdo_line, linewidth = 0.5) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Blind Prediction: PDO + ENSO Only (No SSTs)") +
  theme_paashi()

p_blind_sst / p_blind_pdo
save_fig(p_blind_sst / p_blind_pdo, "Figure_S_Blind_SST_vs_PDO.tiff", height_mm = 140)

# SST+ARMA follows multi-century mean; PDO+ENSO tracks decadal swings in the predicted mean. The structure visible in the PDO+ENSO panel is what our SST-based hindcast CIs represent.

## PDO is strongly related to lake elevation (+0.79m per unit [0.68, 0.91]), ENSO adds +0.20m per unit [0.16, 0.25]. LOO confirms PDO+ENSO improves fit substantially over SSTs alone (~122 elpd). Blind predictions show that PDO+ENSO produces year-to-year and decadal variation in the predicted mean, because it is annually-resolved Pacific climate data rather than a multi-century average.

# NOTE, these PDO and ENSO records only cover ~1,000 years, so they cannot be used for the 23,000-year hindcast. The year-to-year and decadal variability they explain is what our hindcast prediction intervals represent.

# Residual periodograms: does the ~40yr peak survive after removing climate forcing?
# Panel A: SSTs + ARMA(1,1) residuals (from mld_Longarma, full training period)
arma_resids <- residuals(mld_Longarma, type = "ordinary")[, "Estimate"]
sp_resid <- spec.pgram(arma_resids, spans = c(7, 7), plot = FALSE)
sp_resid_period <- 1 / sp_resid$freq
sp_resid_mask <- sp_resid_period >= 10 & sp_resid_period <= 200
df_sp_sst <- data.frame(period = sp_resid_period[sp_resid_mask],
                        spectrum = sp_resid$spec[sp_resid_mask])

fig_periodo_sst <- ggplot(df_sp_sst, aes(x = period, y = spectrum)) +
  geom_line(color = col_model_line, linewidth = 0.5) +
  labs(x = "Period (years)", y = "Spectrum",
       title = "A. Residual Periodogram: SSTs + ARMA(1,1)") +
  theme_paashi()

# Panel B: PDO + ENSO + ARMA(1,1) residuals (~1,000yr subset)
pdo_resids <- residuals(mld_test_pdo_enso_only, type = "ordinary")[, "Estimate"]
sp_pdo <- spec.pgram(pdo_resids, spans = c(7, 7), plot = FALSE)
sp_pdo_period <- 1 / sp_pdo$freq
sp_pdo_mask <- sp_pdo_period >= 10 & sp_pdo_period <= 200

df_sp_pdo <- data.frame(period = sp_pdo_period[sp_pdo_mask],
                        spectrum = sp_pdo$spec[sp_pdo_mask])

fig_periodo_pdo <- ggplot(df_sp_pdo, aes(x = period, y = spectrum)) +
  geom_line(color = col_pdo_line, linewidth = 0.5) +
  labs(x = "Period (years)", y = "Spectrum",
       title = "B. Residual Periodogram: PDO + ENSO + ARMA(1,1)") +
  theme_paashi()

fig_periodograms <- fig_periodo_sst / fig_periodo_pdo
fig_periodograms
save_fig(fig_periodograms, "Figure_S_Residual_Periodograms.tiff", height_mm = 140)

```

## Physical Origin of Periodicity- Examine Adams (2015) Hydrological Model
```{r adams_physics_sims}
#So, the periodicity that survives in our residuals after removing SSTs + ARMA(1,1) is ALSO not a climate signal, as it remains even after using PDO and ENSO. We can test whether it is a physical property of the lake basin. Adams (2015) modeled Tulare Lake as a terminal basin where water only leaves via evaporation. He also notes ~32-64 year periodicity, and suggests that it is a combination of climate signals filtered through the lake's unique hydrology. 

#This idea makes  a lot of sense because surface area scales nonlinearly with volume, so linear increases in water volume or elevation create exponential increases in surface area, massively increasing evaporation. large lake ->  high evaporation -> lake shrinks -> less surface area -> less evaporation -> lake grows. This feedback likely produces the ~32-64 year oscillations regardless of the climate forcing pattern.

#To test this, we use a Bayesian model to simulate the INPUTS of Adams' model, river discharge (Q_in) and direct precipitation (precip). The simulated data should have the same mean and variance as the real input data, but will not follow the exact pattern that the climate did in reality. So, when we then we run these simulated draws through Adams' water balance equations, the periodicity should emerge from the physics alone.

#Note: because we are feeding predictions through a deterministic physics model, we cannot look directly at residual periodograms. We also cannot use the results of model fitting to relate input variables directly to our observed elev variable, since it is processed through another layer of equations. So, this is useful for understanding the physics of the basin, but not for the needs of our hindcast.

# Set up Adams (2015) water balance model from shared Excel spreadsheet (some help from Claude Code)
# Adams (2015) water balance data (extracted from his spreadsheet; cite Adams 2015)
adams_raw <- read.csv("Raw Data/Adams_WB_Data.csv")

# Merge Adams data with climate data (clim already loaded)
wb <- adams_raw %>%
  mutate(Calendar_Year = as.numeric(Year)) %>%
  left_join(clim, by = "Calendar_Year")
wb$Precip_cm <- wb$Precip_km * 1e5  # scale to cm/yr because models failed to converge with wildly different number sizes of predicted variables

# Area (km2) from volume (km3) — 4th order polynomial (Adams 2015)
adams_area <- function(V) {
  -0.2772258143 * V^4 + 8.5611176853 * V^3 -
    89.7086376395 * V^2 + 495.4302443649 * V + 148.4974677424
}

# Elevation (m asl) from volume (km3) — cubic (Adams 2015)
adams_elev <- function(V) {
  0.0058131042 * V^3 - 0.1668677754 * V^2 + 2.0348749694 * V + 56.3610923901
}

# Run Adams' water balance forward for a single draw
run_adams_wb <- function(Q_in_series, Precip_series = NULL, V_init = 1.98,
                         evap_range = c(145, 155)) {
  n <- length(Q_in_series)
  V <- numeric(n)
  V_prev <- V_init
  for (t in seq_len(n)) {
    A <- adams_area(V_prev)
    A <- max(A, 100)
    E <- runif(1, evap_range[1], evap_range[2]) / 100000  # cm/yr -> km/yr
    P <- if (!is.null(Precip_series)) Precip_series[t] else 0
    dV <- Q_in_series[t] + P * A - E * A
    V[t] <- max(V_prev + dV, 0.01)
    V_prev <- V[t]
  }
  adams_elev(V)
}


# Multivariate model: jointly predict Q_in and Precip from PDO + ENSO
wb_pdo <- wb %>%
  filter(!is.na(Q_in), !is.na(Precip_cm), !is.na(PDO), !is.na(ENSO))

model_path <- "Models/mv_pdo.rds"
if (file.exists(model_path)) {
  mv_pdo <- readRDS(model_path)
} else {

mv_pdo <- brm(bf(Q_in ~ PDO + ENSO) +
              bf(Precip_cm ~ PDO + ENSO) +
              set_rescor(TRUE), #this tells a multivariate model that the variables being predicted are correlated
              data = wb_pdo,
              family = gaussian(),
              prior = c(prior(normal(1.5, 1), class = "Intercept", resp = "Qin"),
                        prior(normal(0, 1), class = "b", resp = "Qin"),
                        prior(student_t(3, 0, 1), class = "sigma", resp = "Qin"),
                        prior(normal(20, 10), class = "Intercept", resp = "Precipcm"),
                        prior(normal(0, 10), class = "b", resp = "Precipcm"),
                        prior(student_t(3, 0, 5), class = "sigma", resp = "Precipcm"),
                        prior(lkj(2), class = "rescor")),
              chains = 4,
              iter = 2000,
              warmup = 1000,
              seed = 2025)

  saveRDS(mv_pdo, file = model_path)
}

summary(mv_pdo)


# Posterior predictions. Using the same draws preserves Q_in and Precip correlation
qin_pred_pdo <- posterior_predict(mv_pdo, newdata = wb_pdo, resp = "Qin")
precip_pred_pdo_cm <- posterior_predict(mv_pdo, newdata = wb_pdo, resp = "Precipcm")
qin_pred_pdo[qin_pred_pdo < 0.01] <- 0.01
precip_pred_pdo_cm[precip_pred_pdo_cm < 0] <- 0
precip_pred_pdo <- precip_pred_pdo_cm / 1e5  # back to km/yr for Adams physics

# Run 500 draws through Adams' water balance
n_draws_wb <- 500
set.seed(2025)
draw_idx_wb <- sample(nrow(qin_pred_pdo), n_draws_wb)

years_wb <- wb_pdo$Calendar_Year
obs_wb <- wb_pdo %>% select(Calendar_Year, elev) %>% filter(!is.na(elev))

elev_sim_wb <- matrix(NA, nrow = n_draws_wb, ncol = ncol(qin_pred_pdo))
for (i in seq_len(n_draws_wb)) {
  elev_sim_wb[i, ] <- run_adams_wb(qin_pred_pdo[draw_idx_wb[i], ],
                                   precip_pred_pdo[draw_idx_wb[i], ])
}

## Make a plot to visualize results.
sim_long_wb <- as.data.frame(elev_sim_wb) %>%
  mutate(draw = row_number()) %>%
  pivot_longer(-draw, names_to = "year_idx", values_to = "elev_sim") %>%
  mutate(year_idx = as.integer(gsub("V", "", year_idx)),
         Calendar_Year = years_wb[year_idx])

ggplot() +
  geom_line(data = sim_long_wb,
            aes(x = Calendar_Year, y = elev_sim, group = draw),
            color = col_wb_draw, alpha = 0.01, linewidth = 0.3) +
  geom_line(data = obs_wb,
            aes(x = Calendar_Year, y = elev),
            color = col_observed, linewidth = 0.4) +
  labs(x = "Calendar Year", y = "Water Surface Elevation (m asl)",
       title = "Water Balance Simulations vs Observed",
       subtitle = paste0(n_draws_wb,
                         " posterior draws: (Q_in, Precip) ~ PDO + ENSO -> Adams physics -> elevation")) +
  theme_paashi()

#Look at periodograms (Assisted by Claude Code)
obs_wb_ts <- ts(obs_wb$elev, frequency = 1)
obs_wb_spec <- spectrum(obs_wb_ts, spans = c(5, 7), plot = FALSE)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(1/obs_wb_spec$freq, obs_wb_spec$spec, type = "l", log = "xy",
     xlab = "Period (years)", ylab = "Power",
     main = "Observed elevation", col = col_observed, lwd = 1.5)

for (j in 1:3) {
  draw_ts <- ts(elev_sim_wb[j, ], frequency = 1)
  draw_spec <- spectrum(draw_ts, spans = c(5, 7), plot = FALSE)
  plot(1/draw_spec$freq, draw_spec$spec, type = "l", log = "xy",
       xlab = "Period (years)", ylab = "Power",
       main = paste("WB sim draw", j), col = col_wb_draw, lwd = 1.5)
}

# The simulated periodograms show smooth bumps in the 32-64yr band. This confirms the periodicity is an emergent property of the basin's Area-Volume feedback, not just a feature of the climate forcing

## For our hindcast, we want to directly predict the lake's end state from SSTs, not obscure it through physics, so we use ARMA to help approximate the persistence in lake elevations, which helps inform our CI width. The periodicity inherent to the lake's physics is contained within the CIs of our hindcast.

```

## Using Models to Predict Reality
```{r predictions}
## Now that we understand what our model can tell us and what our CIs represent, let's do an out of sample prediction using Longarma.
#there are several ways to do a posterior prediction. posterior_predict() runs a full simulation based on a trained model, with thousands of draws, outputting every result. This is usually used to compare a model to actual observed results. predict() runs posterior_predict, but then just outputs the mean and upper/lower 95; it is most useful if you just want to go straight to plotting results. posterior_epred ("expected prediction") runs a simulation based on a trained model, gets the mean, then makes predictions based on the mean. This is typically used to make posterior predictions on large sets of new data, since it is significantly lighter than running a full posterior_predict, given that posterior_epred() is just relying on the mean of an already trained model, whereas posterior_predict() or predict() would run thousands of predictions on all the new data, then return the mean.Lastly, fitted() runs posterior_epred() and then summarizes it, much as predict does for posterior_predict.

#First, let's use predict() to visualize how good the model is when arma has observations to work with. This will not have new data, it is just to compare the model against observations

as.data.frame(predict(mld_Longarma)) -> Longarma_Predict #outputs a data frame

#add time and observed data
Longarma_Predict$year_bp <- mld_Longarma$data$Year_B2005
Longarma_Predict$elev_obs <- mld_Longarma$data$elev

#create a csv that we can load later for plotting, rather than running this again
write.csv(Longarma_Predict, "Processed Data/Longarma_Predict.csv", row.names = FALSE)
#for later:
read.csv("Processed Data/Longarma_Predict.csv") -> Longarma_Predict

##Let's visualize the results quickly. We will do more in depth plotting later
#plot light blue ribbon of lower and upper 95, dark blue line is mean, and gray "highlight" is actual values
ggplot(Longarma_Predict, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "skyblue", alpha = 0.5) +
  geom_line(aes(y = Estimate), color = "blue", linewidth = 1) +
  geom_line(aes(y = elev_obs), color = "black", linewidth = 1.5, alpha = 0.4) +
  labs(x = "Years Before Present", y = "Lake Elevation (MaSL)",title = "Observed vs. Modeled Lake Elevation") +
  theme_minimal()

##now let's run it back through time with "new data" from the rest of the climate dataset

#Build clean "new data" for prediction
newdata_pred <- clim %>%
  select(Year_B2005, Seki_Int, Palmer_Int, Davis5, elev) %>%
  filter(!is.na(Year_B2005), !is.na(Seki_Int), !is.na(Palmer_Int), !is.na(Davis5)) %>%
  arrange(Year_B2005)

# fitted() gives the expected mean trend with uncertainty about where the average is.
# predict() gives individual year prediction intervals with noise.

# Expected trend (mean uncertainty only)
Longarma_Fitted <- as.data.frame(fitted(mld_Longarma, newdata = newdata_pred))
colnames(Longarma_Fitted) <- c("Fit_Est", "Fit_SE", "Fit_Q2.5", "Fit_Q97.5")
Longarma_Fitted$year_bp <- newdata_pred$Year_B2005

# Full prediction intervals, uncertainty + variability
Longarma_FullPred <- as.data.frame(predict(mld_Longarma, newdata = newdata_pred))
colnames(Longarma_FullPred) <- c("Pred_Est", "Pred_SE", "Pred_Q2.5", "Pred_Q97.5")
Longarma_FullPred$year_bp <- newdata_pred$Year_B2005

# Combine into one data frame for plotting
hindcast <- left_join(Longarma_Fitted, Longarma_FullPred, by = "year_bp")
hindcast$elev_obs <- newdata_pred$elev

write.csv(hindcast, "Processed Data/Hindcast_Full.csv", row.names = FALSE)
hindcast <- read.csv("Processed Data/Hindcast_Full.csv")

# two-ribbon plot: outer = prediction interval, inner = credible interval for mean
ggplot(hindcast, aes(x = year_bp)) +
  geom_ribbon(aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5),
              fill = col_model_ribbon, alpha = 0.3) +
  geom_ribbon(aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5),
              fill = col_model_inner, alpha = 0.5) +
  geom_line(aes(y = Fit_Est), color = col_model_line, linewidth = 0.5) +
  geom_line(aes(y = elev_obs), color = col_observed, linewidth = 0.5,
            alpha = 0.7, na.rm = TRUE) +
  labs(x = "Years Before 2005", y = "Lake Elevation (MaSL)",
       title = "Lake Elevation Hindcast: Expected Trend and Prediction Intervals",
       subtitle = "Inner band (dark) = 95% CI for mean; Outer band (light) = 95% prediction interval; Dark = observed") +
  theme_paashi()

```

## Publication Figures

### Full Hindcast Figure (assisted by Claude Code)
```{r fig4_hindcast, fig.width = 7.5, fig.height = 5}
hindcast <- read.csv("Processed Data/Hindcast_Full.csv")
training_boundary <- 2006
pleis_holo        <- 11700

# Panel A: Holocene (0-12,000 yr BP)
hindcast_holo <- hindcast %>% filter(year_bp <= 12000)
hindcast_holo_model <- hindcast_holo %>% filter(year_bp > training_boundary)
hindcast_model <- hindcast %>% filter(year_bp > training_boundary)

fig4a <- ggplot(hindcast_holo, aes(x = year_bp)) +
  annotate("rect", xmin = 0, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(data = hindcast_holo_model,
              aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5, fill = "95% prediction interval"), alpha = 0.4) +
  geom_ribbon(data = hindcast_holo_model,
              aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5, fill = "95% CI (expected trend)"), alpha = 0.5) +
  geom_line(data = hindcast_holo_model,
            aes(y = Fit_Est, color = "Model trend"), linewidth = 0.4) +
  geom_line(aes(y = elev_obs, color = "Observed"), linewidth = 0.3,
            alpha = 0.7, na.rm = TRUE) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.3) +
  annotate("text", x = training_boundary + 200, y = 69,
           label = "Training
boundary", size = 2.2, hjust = 0, color = "gray40") +
  annotate("text", x = pleis_holo - 200, y = 69,
           label = "Pleistocene-
Holocene
boundary", size = 2.2,
           hjust = 1, color = col_pleis_line) +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_ribbon,
               "95% CI (expected trend)" = col_model_inner)) +
  scale_color_manual(name = NULL,
    values = c("Model trend" = col_model_line,
               "Observed" = col_observed)) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 12000, by = 2000)) +
  coord_cartesian(ylim = c(50, 72)) +
  labs(x = NULL, y = "Lake Elevation (m asl)",
       title = "A. Holocene (0–12,000 yr BP)") +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme_paashi() +
  theme(legend.position = "none")

fig4b <- ggplot(hindcast, aes(x = year_bp)) +
  annotate("rect", xmin = 0, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  geom_ribbon(data = hindcast_model,
              aes(ymin = Pred_Q2.5, ymax = Pred_Q97.5, fill = "95% prediction interval"), alpha = 0.4) +
  geom_ribbon(data = hindcast_model,
              aes(ymin = Fit_Q2.5, ymax = Fit_Q97.5, fill = "95% CI (expected trend)"), alpha = 0.5) +
  geom_line(data = hindcast_model,
            aes(y = Fit_Est, color = "Model trend"), linewidth = 0.4) +
  geom_line(aes(y = elev_obs, color = "Observed"), linewidth = 0.3,
            alpha = 0.7, na.rm = TRUE) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.3) +
  annotate("text", x = 18000, y = 70,
           label = "Model trained on Holocene data;
deeper predictions carry
greater uncertainty",
           size = 2.0, color = "gray40", fontface = "italic") +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_ribbon,
               "95% CI (expected trend)" = col_model_inner)) +
  scale_color_manual(name = NULL,
    values = c("Model trend" = col_model_line,
               "Observed" = col_observed)) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 22000, by = 4000)) +
  coord_cartesian(ylim = c(50, 72)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Lake Elevation (m asl)",
       title = "B. Full reconstruction (0–23,000 yr BP)") +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme_paashi() +
  theme(legend.position = "bottom")

fig4 <- fig4a / fig4b +
  plot_annotation(
    title = "Bayesian Hindcast of Tulare Lake Elevation")
fig4
save_fig(fig4, "Figure_4_Hindcast.tiff", width_mm = 190, height_mm = 160)
```

### Archaeological Reconstruction Comparison (Assisted by Claude Code)
```{r fig5_arch_comparison, fig.width = 7.5, fig.height = 3.5}
combo <- read.csv("Processed Data/Model_Outputs_Combined.csv")
arch  <- read.csv("Processed Data/Lake12All.csv")
# hindcast has full prediction intervals; combo only has them for the training period
hindcast_holo <- hindcast %>% filter(year_bp <= 12000)
hindcast_holo_model <- hindcast_holo %>% filter(year_bp > training_boundary)
combo_holo <- combo %>% filter(year_b2005 <= 12000)
combo_holo_model <- combo_holo %>% filter(year_b2005 > training_boundary)
arch_holo  <- arch %>% filter(ybp <= 12000)

fig5 <- ggplot() +
  annotate("rect", xmin = 0, xmax = training_boundary,
           ymin = -Inf, ymax = Inf, fill = col_train_bg, alpha = 0.6) +
  # Archaeological raw evidence
  geom_point(data = arch_holo, aes(x = ybp, y = elev),
             color = "gray75", size = 0.1, alpha = 0.3) +
  # Full prediction interval from hindcast data (outside training only)
  geom_ribbon(data = hindcast_holo_model,
              aes(x = year_bp, ymin = Pred_Q2.5, ymax = Pred_Q97.5,
                  fill = "95% prediction interval"), alpha = 0.5) +
  # Model trend (outside training only)
  geom_line(data = combo_holo_model,
            aes(x = year_b2005, y = Fitted_Estimate,
                color = "Model trend"), linewidth = 0.4, alpha = 0.8) +
  # Archaeological reconstruction
  geom_line(data = combo_holo %>% filter(!is.na(arch_model)),
            aes(x = year_b2005, y = arch_model,
                color = "Archaeological reconstruction"), linewidth = 0.6, alpha = 0.85) +
  # Observed lake elevation
  geom_line(data = combo_holo %>% filter(!is.na(elev_obs)),
            aes(x = year_b2005, y = elev_obs,
                color = "Observed"), linewidth = 0.25, alpha = 0.5) +
  geom_vline(xintercept = training_boundary, linetype = "dashed",
             color = "gray40", linewidth = 0.3) +
  geom_vline(xintercept = pleis_holo, linetype = "dotted",
             color = col_pleis_line, linewidth = 0.3) +
  annotate("text", x = pleis_holo - 200, y = 71,
           label = "Pleistocene-
Holocene boundary",
           size = 2.0, hjust = 1, color = col_pleis_line) +
  scale_fill_manual(name = NULL,
    values = c("95% prediction interval" = col_model_ribbon)) +
  scale_color_manual(name = NULL,
    values = c("Model trend" = col_model_line,
               "Archaeological reconstruction" = col_arch,
               "Observed" = col_observed)) +
  scale_x_continuous(labels = comma_format(),
                     breaks = seq(0, 12000, by = 2000)) +
  coord_cartesian(ylim = c(48, 73)) +
  labs(x = "Years Before Present (cal yr BP)",
       y = "Lake Elevation (m asl)",
       title = "Model Reconstruction vs. Archaeological Evidence") +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme_paashi() +
  theme(legend.position = "bottom")
fig5
save_fig(fig5, "Figure_5_Arch_Comparison.tiff", width_mm = 190, height_mm = 100)
```

## Session Info
```{r session_info}
sessionInfo()
```
